---
title: "ACS variables in CHRR"
format:
  html:
    author: Anton Babkin
    date: today
    toc: true
    toc-depth: 2
    embed-resources: true
    code-fold: true
    code-overflow: wrap
    df-print: kable
---


Pilot the calculation of Homeownership (v153), Severe Housing Problems (v136), Some College (v069), High School Completion (v168), Residential Segregation – Black/White (v141), Disconnected Youth (v149), % Not Proficient in English (v059), Children in Single-Parent Households (v082), Driving Alone to Work (v067), Long Commute – Driving Alone (v137)

# References

- [Links to ACS-5 data](https://www.census.gov/data/developers/data-sets/acs-5year.html)
- [Annotation varaibles](https://www.census.gov/data/developers/data-sets/acs-1year/notes-on-acs-estimate-and-annotation-values.html)

# Completion status

Calculated measures from ACS 2023 are compared against previously calculated CHRR-2025 values.
Old calculations did not apply "max variance" rule for zero-valued estimates.

- % Not Proficient in English (v059)
  - Complete, numerically equal

- Driving Alone to Work (v067)
  - Base measure complete, numerically equal
  - Calc by race not implemented

- Some College (v069)
  - Complete, numerically equal

- Children in Single-Parent Households (v082)
  - Complete, numerically equal

- Severe Housing Problems (v136)
  - Base variable complete, numerically equal except:
    - CT, needs crosswalk from old to new via tracts
    - FIPS 02164, CI NA?

- Long Commute – Driving Alone (v137)
  - Complete, numerically equal
  - No ratio calculation, proportions already in raw data.
  - Wrong denominator value in source code (`denominator = S0802_C01_001E`, total)? Should be `S0802_C02_001E`, drove alone.

- Residential Segregation – Black/White (v141)
  - Complete, numerically equal except:
    - US and 2 NY counties (Suffolk County 36103, Ulster County  36111)

- Disconnected Youth (v149)
  - Complete, numerically equal
  - Special case handling, maybe include in general function?

- Homeownership (v153)
  - Complete, numerically equal

- High School Completion (v168)
  - Complete, numerically equal


# Libraries

```{r}
#| output: false

library(tidyverse)

source("R/acs.R")

ACS_YEAR <- 2023
CHAS_YEAR <- 2021

```


# Data

Full tables are pulled from Census Data API and saved to local cache for repeated access.

```{r}
#| eval: false
get_raw_table(ACS_YEAR, "B01001") %>%
  head()

```

## cache

Pack cached raw input files to Zip.

```{r}
#| eval: false
# list of all raw data tables used in the calculations
c(
  str_glue(opath$acs5_vars_, year = ACS_YEAR, tab_type = c("B", "S")),
  str_glue(opath$acs5_table_, year = ACS_YEAR, tract = "",
    tabid = c("B16005", "B08301", "B15001", "B09005", "S0802", "B14005", "B25003", "S1501")),
  str_glue(opath$acs5_table_, year = ACS_YEAR, tract = "_tract",
    tabid = c("B11002A", "B11002B")),
  str_glue(opath$chas_, years = paste0(CHAS_YEAR-4, "thru", CHAS_YEAR), geo = c("040", "050"))
) %>%
  util$zip_pack("acs_raw_inputs_2026-06-03.zip", files = ., overwrite = TRUE)

```


Unpack raw input files from Zip.

```{r}
#| eval: false
util$zip_unpack("acs_raw_inputs_2026-06-03.zip", overwrite = TRUE)

```


## ACS variable metadata

```{r}
VAR_META <- list(
  BC = get_raw_vars(ACS_YEAR, tab_type = "B"),
  S = get_raw_vars(ACS_YEAR, tab_type = "S")
) %>%
  bind_rows()

#' Metadata for variables of given expressions
var_meta_table <- function(...) {
  var_expressions <- list(...)
  var_expressions %>%
    map(\(vex) {
      VAR_META %>%
        filter(variable %in% all.vars(vex)) %>%
        arrange(variable)
    }) %>%
    bind_rows(.id = "term") %>%
    select(term, variable, label)
}


```


# v059 - % Not Proficient in English


Numerator:

The numerator is the number of people aged 5 and over who self-reported they are not proficient in speaking English (speak English either “not well” or “not at all”).

Denominator:

The denominator is the total county population aged 5 and over.


```{r}
tabid <- "B16005"
numerator <- expr(B16005_007E + B16005_012E + B16005_008E + B16005_013E + B16005_017E + B16005_022E + B16005_018E + B16005_023E + B16005_029E + B16005_030E + B16005_034E + B16005_035E + B16005_039E + B16005_040E + B16005_044E + B16005_045E)
denominator <- expr(B16005_001E)

cat("National HTML table:", data_census_gov_url(ACS_YEAR, tabid), "\n")
data.frame(term = c("numerator", "denominator"),
  formula = c(deparse1(numerator), deparse1(denominator)))
var_meta_table(numerator = numerator, denominator = denominator)


x <- get_raw_table(ACS_YEAR, tabid) %>%
  standardize_fips() %>% 
  add_flag_CT() %>%
  # calc_acs_ratio(num_expr = !!numerator, den_expr = !!denominator, zero_max = FALSE) %>%
  calc_acs_ratio(num_expr = !!numerator, den_expr = !!denominator) %>%
  apply_suppression() %>%
  mutate(sourceflag = NA) %>%
  add_col_prefix("v059") %>%
  arrange(statecode, countycode)

v059 <- x %>%
  select(statecode, countycode, starts_with("v059"))

skimr::skim(v059)

```

```{r}
#| echo: false
#| eval: false
x0 <- read_csv("measure_datasets/v059_r2025.csv") %>%
  arrange(statecode, countycode)

waldo::compare(x0, v059, tolerance = 1e-4, list_as_map = TRUE)

```


# v067 - PERCENT OF LABOR FORCE THAT DRIVES ALONE TO WORK

numerator =	number of workers who commute alone to work via car, truck, or van.

denominator	= total workforce

raw_value	= Percentage of workers who drive alone to work

```{r}
tabid <- "B08301"
numerator <- expr(B08301_003E)
denominator <- expr(B08301_001E)

cat("National HTML table:", data_census_gov_url(ACS_YEAR, tabid), "\n")
data.frame(term = c("numerator", "denominator"),
  formula = c(deparse1(numerator), deparse1(denominator)))
var_meta_table(numerator = numerator, denominator = denominator)


x <- get_raw_table(ACS_YEAR, tabid) %>%
  standardize_fips() %>% 
  add_flag_CT() %>%
  calc_acs_ratio(num_expr = !!numerator, den_expr = !!denominator) %>%
  apply_suppression() %>%
  mutate(sourceflag = NA) %>%
  add_col_prefix("v067")

v067 <- x %>%
  select(statecode, countycode, starts_with("v067"))

skimr::skim(v067)

```

```{r}
#| echo: false
#| eval: false
x0 <- read_csv("measure_datasets/v067_r2025.csv") %>%
  arrange(statecode, countycode)

waldo::compare(x0, v067, tolerance = 1e-4, list_as_map = TRUE)

```


## v067_other - Driving Alone to Work

NOT IMPLEMENTED

- `race_aian`	AIAN	Percentage of American Indian and Alaska Native workers who drive alone to work	
- `race_asian`	Asian	Percentage of Asian/Pacific Islander workers who drive alone to work
- `race_black`	Black	Percentage of Black workers who drive alone to work	Percentage
- `race_hispanic`	Hispanic	Percentage of Hispanic workers who drive alone to work
- `race_white`	white	Percentage of non-Hispanic white workers who drive alone to work


```{r}
var_ls_067_other <- c(
  'B08505B_002E', 'B08505H_002E', 'B08505I_002E', 'B08505D_002E', 'B08505E_002E',
  'B08505C_002E', 'B08505B_001E', 'B08505H_001E', 'B08505I_001E', 'B08505D_001E',
  'B08505E_001E', 'B08505C_001E', 'B08505B_002M', 'B08505H_002M', 'B08505I_002M',
  'B08505D_002M', 'B08505E_002M', 'B08505C_002M', 'B08505B_001M', 'B08505H_001M',
  'B08505I_001M', 'B08505D_001M', 'B08505E_001M', 'B08505C_001M'
)
```


# v069 - SOME COLLEGE

numerator =	Total number of individuals ages 25-44 with at least some post-secondary education

denominator	= Total number of individuals ages 25-44.

raw_value	= Percentage of adults age 25-44 with some post-secondary education

```{r}
tabid <- "B15001"
numerator <- expr(B15001_015E + B15001_016E + B15001_017E + B15001_018E + B15001_023E + 
        B15001_024E + B15001_025E + B15001_026E + B15001_056E + B15001_057E + B15001_058E +  
        B15001_059E + B15001_064E + B15001_065E + B15001_066E + B15001_067E)
denominator <- expr(B15001_011E + B15001_019E + B15001_052E + B15001_060E)

cat("National HTML table:", data_census_gov_url(ACS_YEAR, tabid), "\n")
data.frame(term = c("numerator", "denominator"),
  formula = c(deparse1(numerator), deparse1(denominator)))
var_meta_table(numerator = numerator, denominator = denominator)


x <- get_raw_table(ACS_YEAR, tabid) %>%
  standardize_fips() %>% 
  add_flag_CT() %>%
  calc_acs_ratio(num_expr = !!numerator, den_expr = !!denominator, zero_max = FALSE) %>%
  # calc_acs_ratio(num_expr = !!numerator, den_expr = !!denominator) %>%
  apply_suppression() %>%
  mutate(sourceflag = NA) %>%
  add_col_prefix("v069")

v069 <- x %>%
  select(statecode, countycode, starts_with("v069"))

skimr::skim(v069)

```


```{r}
#| echo: false
#| eval: false

x0 <- read_csv("measure_datasets/v069_r2025.csv") %>%
  arrange(statecode, countycode)

waldo::compare(x0, v069, tolerance = 1e-4, list_as_map = TRUE)

```


# v082 - SINGLE-PARENT HOUSEHOLDS

numerator =	Number of children that live in single-parent households

denominator	= Number of children in households

raw_value	= Percentage of children that live in single-parent households


```{r}
tabid <- "B09005"
numerator <- expr(B09005_004E + B09005_005E)
denominator <- expr(B09005_001E)

cat("National HTML table:", data_census_gov_url(ACS_YEAR, tabid), "\n")
data.frame(term = c("numerator", "denominator"),
  formula = c(deparse1(numerator), deparse1(denominator)))
var_meta_table(numerator = numerator, denominator = denominator)

x <- get_raw_table(ACS_YEAR, tabid) %>%
  standardize_fips() %>% 
  add_flag_CT() %>%
  calc_acs_ratio(num_expr = !!numerator, den_expr = !!denominator) %>%
  # calc_acs_ratio(num_expr = !!numerator, den_expr = !!denominator, zero_max = FALSE) %>%
  apply_suppression() %>%
  # repeating original edge-case code, might want to generalize to work with any zero values
  # setting '15005' to missing because numerator and denominator are 0 which isn't helpful
  mutate(across(c(cilow, cihigh), ~if_else(is.nan(rawvalue), NA, .))) %>% 
  mutate(numerator = if_else((statecode=="15" & countycode=="005"), NA, numerator),
         denominator = if_else((statecode=="15" & countycode=="005"), NA, denominator),
         cilow = if_else((statecode=="15" & countycode=="005"), NA, cilow),
         cihigh = if_else((statecode=="15" & countycode=="005"), NA, cihigh)) %>% 
  mutate(sourceflag = NA) %>%
  add_col_prefix("v082")

v082 <- x %>%
  select(statecode, countycode, starts_with("v082"))

skimr::skim(v082)

```


```{r}
#| echo: false
#| eval: false

x0 <- read_csv("measure_datasets/v082_r2025.csv") %>%
  arrange(statecode, countycode)

waldo::compare(x0, v082, tolerance = 1e-4, list_as_map = TRUE)

```


# v136 - Severe Housing Problems

Columns are similar to ACS columns, but are named differently.
We rename them to ACS format here to reuse ratio and MOE calculation function.


```{r}

numerator <- expr(T2_3E + T2_76E)
denominator <- expr(T2_1E)


x1 <- bind_rows(
  get_raw_chas(CHAS_YEAR, "050", "Table2"),
  get_raw_chas(CHAS_YEAR, "040", "Table2") %>% mutate(cnty = "000")
) %>%
  rename(ucgid = geoid) %>%
  standardize_fips() %>%
  filter(statecode != "00") %>%
  add_flag_CT() %>%
  rename_with(\(x) {
    ifelse(substr(x, 4, 6) == "est", sub("est(.+)$", "\\1E", x), sub("moe(.+)$", "\\1M", x))
  }, .cols = starts_with("T2_")) %>%
  # calc_acs_ratio(num_expr = !!numerator, den_expr = !!denominator)
  calc_acs_ratio(num_expr = !!numerator, den_expr = !!denominator, zero_max = FALSE)

# calc US aggregate from states
x2 <- x1 %>%
  filter(sumlevel == "040") %>%
  summarize(
    statecode = "00",
    countycode = "000",
    numerator = sum(numerator),
    denominator = sum(denominator),
    rawvalue = numerator / denominator
  ) %>%
  bind_rows(x1) %>%
  arrange(statecode, countycode) %>%
  apply_suppression() %>%
  mutate(sourceflag = NA) %>%
  add_col_prefix("v136")

v136 <- x2 %>%
  select(statecode, countycode, starts_with("v136"))

skimr::skim(v136)

```


```{r}
#| echo: false
#| eval: false
x0 <- read_csv("measure_datasets/v136_r2025.csv") %>%
  arrange(statecode, countycode)

waldo::compare(x0, v136, tolerance = 1e-4, list_as_map = TRUE)

xcmp <- full_join(x0, v136, by = c("statecode", "countycode")) %>%
  # known difference CT
  filter(statecode != "09")

x <- xcmp %>%
  # filter(!util$almost_equal(v136_rawvalue.x, v136_rawvalue.y))
  filter(!util$almost_equal(v136_cihigh.x, v136_cihigh.y))
x

```

# v137 - LONG COMMUTE - DRIVING ALONE 

numerator =	number of workers who drive alone (via car, truck, or van) for more than 30 minutes during their commute

denominator	= number of workers who drive alone (via car, truck, or van) during their commute

raw_value	= Among workers who commute in their car alone, the percentage that commute more than 30 minutes


```{r}
tabid <- "S0802"
denominator <- expr(S0802_C02_001E)
rawvalue <- expr(S0802_C02_086E + S0802_C02_087E + S0802_C02_088E + S0802_C02_089E)

cat("National HTML table:", data_census_gov_url(ACS_YEAR, tabid), "\n")
data.frame(term = c("rawvalue", "denominator"),
  formula = c(deparse1(rawvalue), deparse1(denominator)))
var_meta_table(rawvalue = rawvalue, denominator = denominator)

x <- get_raw_table(ACS_YEAR, tabid) %>%
  standardize_fips() %>% 
  add_flag_CT() %>%
  # no ratio, using general function to handle variance calc
  # calc_acs_ratio(num_expr = !!rawvalue, den_expr = !!denominator, zero_max = FALSE) %>%
  calc_acs_ratio(num_expr = !!rawvalue, den_expr = !!denominator) %>%
  mutate(
    rawvalue = numerator / 100,
    numerator = NA,
    var = var_num / 10000,
    var_num = NA,
    var_prop = NA,
    moe = sqrt(var) * 1.96,
    cilow = rawvalue - moe, 
    cihigh = rawvalue + moe
  ) %>%
  apply_suppression() %>%
  mutate(sourceflag = NA) %>%
  add_col_prefix("v137")

v137 <- x %>%
  select(statecode, countycode, starts_with("v137"))

skimr::skim(v137)

```


```{r}
#| echo: false
#| eval: false

x0 <- read_csv("measure_datasets/v137_r2025.csv") %>%
  arrange(statecode, countycode) %>%
  # old uses wrong denominator
  select(-v137_denominator)

waldo::compare(x0, v137, tolerance = 1e-4, list_as_map = TRUE)

```




# v141 - Residential Segregation – Black/White

Index of dissimilarity where higher values indicate greater residential segregation between Black and White county residents.

Calculated with custom code that is different from all other measures.

$$
D_r = \frac{1}{2}\sum_{i \in I_r} 100 \left\lvert \frac{w_i}{W_r} - \frac{b_i}{B_r} \right\rvert
$$

- $D_r$ Dissimilarity index for state or county $r$
- $w_i$ and $b_i$ number of White and Black households in tract $i$
- $W_r$ and $W_i$ number of White and Black households in state or county $r$
- $I_r$ set of tracts in state of county $r$

```{r}

tabid <- c("B11002A", "B11002B")
cat("National HTML table:", data_census_gov_url(ACS_YEAR, tabid[1]), "\n")
cat("National HTML table:", data_census_gov_url(ACS_YEAR, tabid[2]), "\n")
var_meta_table(w = expr(B11002A_001E), b = expr(B11002B_001E))

# Join white and black tables by GEOID
x <- full_join(
    get_raw_table(ACS_YEAR, tabid[1], tract = TRUE) %>% select(NAME, ucgid, wht = B11002A_001E),
    get_raw_table(ACS_YEAR, tabid[2], tract = TRUE) %>% select(ucgid, blk = B11002B_001E),
    by = "ucgid"
  ) %>%
  standardize_fips() %>%
  # only keep tract-level observations
  filter(str_sub(ucgid, 1, 3) == "140")

# calculate dissimilarity index for a groupped dataframe (by state or by county)
calc_diss_index <- function(df) {
  df %>%
    # suppress counties with less than 100 black population
    filter(sum(blk) >= 100) %>%
    mutate(
      gr_wht = sum(wht), # total white population in geographic group
      gr_blk = sum(blk), # total black population in geographic group
      tr_diss = abs(wht / gr_wht - blk / gr_blk) # dissimilarity in tract shares of white and black 
    ) %>%
    # index = (1/2) SUM|(b/B) - (w/W)| * 100)
    summarize(
      diss_index = 50 * sum(tr_diss, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(diss_index = replace_values(diss_index, 0 ~ NA))
}

# apply aggregated calculation at US, state and county level
v141 <- bind_rows(
  x %>%
    group_by(statecode, countycode) %>%
    calc_diss_index(),
  x %>%
    group_by(statecode) %>%
    calc_diss_index() %>%
    mutate(countycode = "000"),
  x %>%
    calc_diss_index() %>%
    mutate(statecode = "00", countycode = "000")
  ) %>%
  arrange(statecode, countycode) %>%
  rename(v141_rawvalue = diss_index)

skimr::skim(v141)

```


```{r}
#| eval: false
#| echo: false

x0 <- read_csv("measure_datasets/v141_r2025.csv") %>%
  arrange(statecode, countycode)

full_join(x0, v141, by = c("statecode", "countycode")) %>%
  filter(!util$almost_equal(v141_rawvalue.x, v141_rawvalue.y))

```


# v149 - DISCONNECTED YOUTH

numerator =	number of people, ages 16-19, who are neither working nor in school

denominator	= total county population, ages 16-19

raw_value	= Percentage of teens and young adults ages 16-19 who are neither working nor in school


```{r}
tabid <- "B14005"
numerator <- expr(B14005_010E + B14005_011E + B14005_014E + B14005_015E +
           B14005_024E + B14005_025E + B14005_028E + B14005_029E)
denominator <- expr(B14005_001E)

cat("National HTML table:", data_census_gov_url(ACS_YEAR, tabid), "\n")
data.frame(term = c("numerator", "denominator"),
  formula = c(deparse1(numerator), deparse1(denominator)))
var_meta_table(numerator = numerator, denominator = denominator)


x <- get_raw_table(ACS_YEAR, tabid) %>%
  standardize_fips() %>% 
  add_flag_CT() %>%
  # calc_acs_ratio(num_expr = !!numerator, den_expr = !!denominator, zero_max = FALSE) %>%
  calc_acs_ratio(num_expr = !!numerator, den_expr = !!denominator) %>%
  apply_suppression() %>%
  mutate(sourceflag = NA) %>%
  # could go inside apply_suppresion?
  mutate(
    rawvalue = case_when(
      rawvalue == 0 ~ NA_real_,
      rawvalue == 1 ~ NA_real_,
      # standard error too high
      sqrt(var) / rawvalue > 0.35 ~ NA_real_,
      TRUE ~ rawvalue
    )
  ) %>%
  mutate(across(c(numerator, denominator, cilow, cihigh), 
                ~if_else(is.na(rawvalue), NA_real_, .x))) %>% 
  add_col_prefix("v149")

v149 <- x %>%
  select(statecode, countycode, starts_with("v149"))

skimr::skim(v149)

```


```{r}
#| echo: false
#| eval: false

x0 <- read_csv("measure_datasets/v149_r2025.csv") %>%
  arrange(statecode, countycode)

waldo::compare(x0, v149, tolerance = 1e-4, list_as_map = TRUE)

```


# v153 - Homeownership

Description: Percentage of owner-occupied housing units.

Numerator: The numerator is the total number of owner-occupied housing units in a county.

Denominator: The denominator is the total occupied housing units in a county.


```{r}
tabid <- "B25003"
numerator <- expr(B25003_002E)
denominator <- expr(B25003_001E)

cat("National HTML table:", data_census_gov_url(ACS_YEAR, tabid), "\n")
data.frame(term = c("numerator", "denominator"),
  formula = c(deparse1(numerator), deparse1(denominator)))
var_meta_table(numerator = numerator, denominator = denominator)


x <- get_raw_table(ACS_YEAR, tabid) %>%
  standardize_fips() %>% 
  add_flag_CT() %>%
  calc_acs_ratio(num_expr = !!numerator, den_expr = !!denominator) %>%
  apply_suppression() %>%
  mutate(sourceflag = NA) %>%
  add_col_prefix("v153")

v153 <- x %>%
  select(statecode, countycode, starts_with("v153"))

skimr::skim(v153)

```


```{r}
#| echo: false
#| eval: false

x0 <- read_csv("measure_datasets/v153_r2025.csv") %>%
  arrange(statecode, countycode)

waldo::compare(x0, v153, tolerance = 1e-4, list_as_map = TRUE)

```


# v168 - HIGH SCHOOL COMPLETION

numerator =	# Completed High School (Adults age 25 and over with a high school diploma or equivalent)

denominator =	Population	(Adults age 25 and over)

raw_value	= % Completed High School


```{r}
tabid <- "S1501"
numerator <- expr(S1501_C01_014E)
denominator <- expr(S1501_C01_006E)

cat("National HTML table:", data_census_gov_url(ACS_YEAR, tabid), "\n")
data.frame(term = c("numerator", "denominator"),
  formula = c(deparse1(numerator), deparse1(denominator)))
var_meta_table(numerator = numerator, denominator = denominator)


x <- get_raw_table(ACS_YEAR, tabid) %>%
  standardize_fips() %>% 
  add_flag_CT() %>%
  calc_acs_ratio(num_expr = !!numerator, den_expr = !!denominator) %>%
  apply_suppression() %>%
  mutate(sourceflag = NA_real_) %>%
  add_col_prefix("v168")

v168 <- x %>%
  select(statecode, countycode, starts_with("v168"))

skimr::skim(v168)

```


```{r}
#| echo: false
#| eval: false

x0 <- read_csv("measure_datasets/v168_r2025.csv") %>%
  arrange(statecode, countycode)

waldo::compare(x0, v168, tolerance = 1e-4, list_as_map = TRUE)

```


# CHRR raw ACS-5

```{r}
#| eval: false
chrr_acs5 <- read_csv("raw_data/ACS/acs_data_2019-2023.zip")%>% 
  rename(fipscode = GEOID) %>% 
  glimpse()
```



# Data diff

Helper code to compare new computations with old CHRR results.

```{r}
#| eval: false
numerator <- expr(S1501_C01_014E)
denominator <- expr(S1501_C01_006E)
chrr_path <- "measure_datasets/v168_r2025.csv"


cmp_vars <- c("statecode", "countycode", "numerator", "denominator", "rawvalue", "cilow", "cihigh")
x0 <- read_csv(chrr_path) %>%
  arrange(statecode, countycode) %>%
  rename_with(\(x) str_remove(x, "^v..._")) %>%
  select(all_of(cmp_vars))

x1 <- get_raw_table(ACS_YEAR, tabid) %>%
  standardize_fips() %>% 
  calc_acs_ratio(num_expr = !!numerator, den_expr = !!denominator, zero_max = FALSE)

cmp <- cmp_vars[-(1:2)] %>%
  set_names() %>%
  map(\(var) {
    waldo::compare(x0[[var]], x1[[var]], tolerance = 1e-4)
  })

cmp


x <- full_join(
  select(x0, statecode, countycode, v0 = cilow),
  select(x1, statecode, countycode, v1 = cilow),
  by = c("statecode", "countycode")
) %>%
  filter(!util$almost_equal(v0, v1))

raw_vars <- base::union(all.vars(numerator), all.vars(denominator)) %>%
  c(., map_chr(., \(xe) sub("E$", "M", xe)))

x1e <- x1 %>%
  right_join(x) %>%
  select(all_of(raw_vars) | !starts_with(tabid))


```

