---
title: "v167"
output: html_document
date: "2025-02-19"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(haven)

data_out_path <- "../01_data/processed"

data_dir_2025 <- 'C:/01_CHRR/data/r2025'
```


# About v167 (School segregation)

The extent to which students of different race/ethnic groups are unevenly distributed across schools when compared with the racial and ethnic composition of the local population. The index ranges from 0 to 1 where a value of zero indicates a school population that directly reflects the community population, while a higher value indicates greater levels of school segregation.

# standard FIPS

```{r}
fips <- bind_rows(
  read_sas(file.path(data_dir_2025, "county_fips.sas7bdat") ),
  read_sas(file.path(data_dir_2025, "state_fips.sas7bdat")))  %>% 
  arrange(statecode, countycode) %>% 
  glimpse()

# with old CT counties
fips_all <- bind_rows(
  fips, 
  read_sas(file.path(data_dir_2025, "county_fips_old.sas7bdat") ) %>% 
    filter(statecode == "09")) %>% 
  arrange(statecode, countycode)
```

# helpers

```{r}
calc_proportations <- function(df){
  res <- df %>% 
    mutate(aian_prop = aian/total_students, 
           asian_prop = asian/total_students,
           hisp_prop = hisp/total_students, 
           black_prop = black/total_students,
           white_prop = white/total_students, 
           two_prop = two/total_students)
  
  return(res)
}

calc_theil <- function(df){
  res <- df %>% 
    mutate(theil = -(    aian_prop  * log(ifelse(aian_prop==0,  1, aian_prop)) 
                       + asian_prop * log(ifelse(asian_prop==0, 1, asian_prop)) 
                       + hisp_prop  * log(ifelse(hisp_prop==0,  1, hisp_prop)) 
                       + black_prop * log(ifelse(black_prop==0, 1, black_prop)) 
                       + white_prop * log(ifelse(white_prop==0, 1, white_prop))
                       + two_prop   * log(ifelse(two_prop==0,   1, two_prop))
                       )
         )
  
  return(res$theil)
}

add_flag_CT <- function(df, vnum, old="U", new = "A"){
  CT_old_lst <- c("001", "003", "005", "007", "009", "011", "013", "015")
  CT_new_lst <- c("110", "120", "130", "140", "150", "160", "170", "180", "190")

  res <- df %>% 
    mutate(flag_CT = case_when(
      statecode == "09" & countycode %in% CT_old_lst ~ old,
      statecode == "09" & countycode %in% CT_new_lst ~ new,
      TRUE ~ NA
    )) %>% 
    rename(!!(paste0(vnum, "_flag_CT")) := flag_CT)
  
  return(res)

}
```


# Import NCES data

-   The actual data starts from line 7 in the raw data set (csv); so set skip = 6 

```{r }
df_nces <- read_csv(
  here::here(data_dir_2025, "raw/nces/school_segregation/ELSI_csv_export_6387151618792713067141.csv"),
  skip = 6,
  n_max = Inf) %>% 
  janitor::clean_names() %>% 
  filter(!is.na(state_name_public_school_latest_available_year)) %>%
  glimpse()

```

```{r}
df_nces %>% head()
```

# clean column names

```{r}
df_nces_clean_name <- df_nces %>% 
  # remove school_name column, which is the same as column 3
  select(-school_name) %>% 
  rename_at(
    # select all variables with "_public_school_2023_24" in the name
    vars(contains("_public_school_2023_24")),
    # use stringr::str_replace to remove _public_school_2023_24
    list(~str_replace(., "_public_school_2023_24", ""))) %>% 
  rename(state = state_name_public_school_latest_available_year,
         fipscode = county_number,
         total_students = total_students_all_grades_excludes_ae,
         aian = american_indian_alaska_native_students,
         asian = asian_or_asian_pacific_islander_students,
         hisp = hispanic_students,
         black = black_or_african_american_students,
         white = white_students,
         nhop = nat_hawaiian_or_other_pacific_isl_students,
         two = two_or_more_races_students) %>% 
  # convert state names to upper case
  mutate(state = str_to_upper(state),
         school_name = str_to_upper(school_name)) %>% 
  glimpse()
```

# explore columns

-   school_type

```{r}
df_nces_clean_name %>% 
  select(school_type) %>% 
  distinct()

```

- start_of_year_status

```{r}
df_nces_clean_name %>% 
  select(start_of_year_status) %>% 
  distinct()
```

```{r}
df_nces_clean_name %>% 
  count(start_of_year_status)
  # filter(start_of_year_status == '6-Inactive') 
```

-   total_students

```{r}
df_nces_clean_name %>% 
  select(total_students) %>% 
  filter(str_detect(total_students, "\\D+|(^0$)")) %>% 
  group_by(total_students) %>% 
  count() %>% 
  ungroup() %>% 
  janitor::adorn_totals("row")
```

-   There are 4348 schools with `total_students` being missing or 0; these schools will be excluded 

# 1. The following school types are excluded

```{r}
df_nces_clean_name %>% 
  distinct(school_type)

df_nces_clean_name %>% 
  distinct(start_of_year_status)
```


-   if `school_name` contains 'JUVENILE|JAIL|PRISON|CORRECTIONAL|CYBER|VIRTUAL' then delete;
-   if `School_Type` == `4-Alternative Education School`, then delete;
-   if `Start_of_Year_Status` == '6-Inactive', then delete;
-   if `Start_of_Year_Status` == '2-Closed', then delete;
-   if `total_students` is missing or 0, then delete

```{r}
df1 <- df_nces_clean_name %>%
  filter(!str_detect(school_name,
                     'JUVENILE|JAIL|PRISON|CORRECTIONAL|CYBER|VIRTUAL')) %>%
  filter(school_type != "4-Alternative Education School") %>%
  filter(start_of_year_status != '6-Inactive') %>%
  filter(start_of_year_status != '2-Closed') %>%
  # filter(start_of_year_status != '7-Future') %>%
  filter(!str_detect(total_students, "\\D+")) %>%  # total_students not 0
  glimpse()
```

# 2. Create proportions for each race/ethnicity category

## clean columns and convert to numeric

```{r}
grps_lst <- c("total_students", "aian", "asian", "hisp", "black", "white", "nhop", "two")

df2a <- df1 %>% 
  mutate(fipscode = str_extract(fipscode, "\\d+")) %>% 
  # get statecode and countycode from fipscode
  mutate(statecode = str_sub(fipscode, 1L, 2L),
         countycode = str_sub(fipscode, 3L, 5L), .after = 'fipscode') %>% 
  mutate(across(all_of(grps_lst),  ~ str_extract(., "\\d+")) ) %>%
  mutate(across(all_of(grps_lst), as.numeric) ) %>%
  mutate(across(all_of(grps_lst), ~ if_else(is.na(.), 0, .)) ) %>%
  mutate(asian = asian + nhop) %>% 
  select(-c(school_name,county_name, school_type, start_of_year_status, nhop)) %>% 
  filter(total_students > 0) %>% # remove schools with 0 total students
  glimpse()
```

```{r}
df2a %>% 
  filter(fipscode == "12065")
```

## calculate proportions

```{r}
df2 <- df2a %>% 
  calc_proportations() %>% 
  glimpse()

```

# 3. Calculate a theil value for each school

theil_school = -(amer_ind_prop*log(amer_ind_prop)+asian_pi_haw_prop*log(asian_pi_haw_prop)+
hispanic_prop*log(hispanic_prop)+black_prop*log(black_prop)+white_prop*log(white_prop)+two_prop*log(two_prop))

$$theil_{school} = -(aian_{prop}*log(aian_{prop})+asian_{prop}*log(asian_{prop})+
hisp_{prop}*log(hisp_{prop})+black_{prop}*log(black_{prop})+white_{prop}*log(white_{prop})+two_{prop}*log(two_{prop}))$$

```{r}
df3 <- df2 %>% 
  mutate(theil_school = calc_theil(.)) %>%
  glimpse()
```

# 4. Calculate a theil value for each county

-   `theil_county` = aggregate school values for each race/ethnicity category and complete the same calculation as above for theil_school

```{r}
df4 <- df2a %>% 
  group_by(fipscode) %>% 
  mutate(across(all_of(grps_lst[grps_lst!='nhop']), sum) ) %>% 
  mutate(school_num = n()) %>%
  distinct(fipscode, .keep_all=TRUE) %>% 
  ungroup() %>% 
  calc_proportations() %>% 
  mutate(theil_county = calc_theil(.)) %>%
  mutate(flag = (aian>=25) + (asian>=25) + (hisp>=25) + (black>=25) + (white>=25)+
           (two>=25) 
         ) %>% 
  glimpse()

```

# 5.Calculate school_weight, theil_school_weight, theil_count_weight

`school_weight` = Total_students_in_school/Total_students_in_county
`theil_school_weight` = theil_school*school_weight
`theil_county_weight` = aggregating the theil_school_weight for each county

```{r}
df5 <- df3 %>% 
  group_by(fipscode) %>% 
  mutate(total_students_cny = sum(total_students, na.rm = TRUE),
         school_weight = total_students/total_students_cny,
         theil_school_weight = theil_school * school_weight
         ) %>% 
  summarise(theil_county_weight = sum(theil_school_weight, na.rm = TRUE)) %>% 
  glimpse()
```

# 6. Create final theil value for each county

`theil_county_final` = abs(`theil_county` - `theil_county_weight`)/`theil_county`

```{r}
df6 <- df4 %>% 
  select(state, fipscode, statecode, countycode, theil_county, flag, school_num) %>% 
  left_join(df5, by = "fipscode") %>% 
  mutate(theil_county_final = abs(theil_county - theil_county_weight)/theil_county) %>% 
  # suppress counties with only one group that has >25 students
  mutate(theil_county_final = if_else(flag < 2, NA_real_, theil_county_final)) %>% 
  # suppress counties with only one school
  mutate(theil_county_final = if_else(school_num < 2, NA_real_, theil_county_final)) %>% 
  glimpse()

```

## check CT: new counties (i.e., palnning regions)

```{r}
df6 %>% 
  filter(statecode == "09")
```

# 7. Calculate a theil value for each state

`theil_state` = aggregate county values for each race/ethnicity category and complete the same calculation as for theil_county

```{r}
df7 <- df2a %>% 
  select(-c(fipscode)) %>% 
  group_by(state) %>% 
  mutate(across(all_of(grps_lst[grps_lst!='nhop']), sum) ) %>%
  distinct(state, .keep_all=TRUE) %>% 
  ungroup() %>%
  calc_proportations() %>% 
  mutate(theil_state = calc_theil(.)) %>%
  glimpse()
```

# 8.Calculate county_weight, theil_county_weight, theil_state_weight

```{r}
# use theil_school
glimpse(df3)
df8 <- df3 %>% 
  select(state, statecode, countycode, total_students, aian, asian, hisp, 
         black, white, two, theil_school) %>% 
  group_by(state) %>% 
  mutate(total_students_st = sum(total_students, na.rm = TRUE),
         school_weight = total_students/total_students_st,
         theil_county_weight = theil_school * school_weight
  ) %>% 
  summarise(theil_state_weight = sum(theil_county_weight, na.rm = TRUE), 
            statecode = first(statecode)) %>% 
  mutate(countycode = '000') %>% 
  glimpse()
```

# 9. Create the final theil value for each state

`theil_state_final` = abs(theil_state-theil_state_weight)/theil_state

```{r}
df9 <- df7 %>% 
  select(state, theil_state) %>% 
  left_join(df8, by = "state") %>% 
  mutate(theil_state_final = abs(theil_state - theil_state_weight)/theil_state) %>% 
  select(state, statecode, countycode, everything()) %>% 
  glimpse()
```

# Combine county and state data

```{r}
df12 <- bind_rows(
  df6 %>% ungroup() %>% 
    select(statecode, countycode, v167_rawvalue = theil_county_final),
  df9 %>% ungroup() %>%
    select(statecode, countycode, v167_rawvalue = theil_state_final)
  ) %>% 
  arrange(statecode, countycode) %>% 
  glimpse()
```

### add us value as median of state values

```{r}
us_rawvalue <- df12 %>% 
  filter(countycode == "000") %>% 
  summarise(median(v167_rawvalue)) %>% 
  pull()

us_rawvalue
```

```{r}
df12b <- df12 %>% 
  add_row(statecode = "00", countycode = "000", v167_rawvalue = us_rawvalue) %>% 
  arrange(statecode, countycode)
```


```{r}
v167_gl <- fips_all %>% 
  left_join(df12b, by = c('statecode', 'countycode')) %>% 
  mutate(v167_numerator = NA_real_, v167_denominator = NA_real_, 
         v167_cilow = NA_real_, v167_cihigh = NA_real_,
         v167_sourceflag = NA_real_) %>% 
  select(-c(fipscode, state, county)) %>% 
  glimpse()
```

## explore

```{r}
summary(v167_gl)
```

# add flag_CT

```{r}
v167_gl_2 <- v167_gl %>% 
  add_flag_CT(vnum = "v167", old="U", new = "A")

v167_gl_2
```

# save data

```{r eval=FALSE}
write_csv(v167_gl_2,
  paste0(
    data_out_path, "/",
    "v167_r2025", 
    "_",  format(Sys.Date(), "%Y_%m_%d"),
    ".csv"
  ),
  na = ""
)
```
