---
title: "v051, v052, v053, v057 - R2026"
format:
  html:
    author: Anton Babkin
    date: today
    toc: true
    toc-depth: 2
    embed-resources: true
    code-fold: true
    code-overflow: wrap
    df-print: kable
---

# About

This notebook calculates the following demographic variables.

- v051, Population
- v052, % Below 18 Years of Age
- v053, % 65 and Older
- v057, % Female


# Import packages 

```{r}
library(tidyverse)
library(haven)

source("calculations/helpers.R", local = (helpers <- new.env()))

# set CHRR_YEAR to 2026 for new r2026 calculations, to 2025 for comparison against old r2025 measures
CHRR_YEAR <- 2025
POPEST_YEAR <- CHRR_YEAR - 2

# set to FALSE to run calculations without saving measures to disk
SAVE_MEASURES <- TRUE

# calculated measures dataframes
measures <- list()

# input paths
ipath <- list(
  agesex_ = "https://www2.census.gov/programs-surveys/popest/datasets/2020-{POPEST_YEAR}/counties/asrh/cc-est{POPEST_YEAR}-agesex-all.csv",
  r2025_measure_ = "measure_datasets/{vname}_r2025.csv"
)

# output paths
opath <- list(
  r2026_measure_ = "measure_datasets/{vname}_r2026.csv"
)

```

# Prepare data


- Website to downloaded data

`cc-est2024-agesex-all.csv`

https://www2.census.gov/programs-surveys/popest/datasets/2020-2024/

- Online documentation

https://www2.census.gov/programs-surveys/popest/technical-documentation/file-layouts/2020-2024/CC-EST2024-AGESEX.pdf



## import Census pop data

data: https://www2.census.gov/programs-surveys/popest/datasets/2020-2024/counties/asrh/cc-est2024-agesex-all.csv

```{r}

df_raw <- str_glue(ipath$agesex_) %>%
  read_csv() %>%
  rename_with(tolower)

df_raw %>%
  head()

```


## process data

POPESTIMATE: Total population
POPEST_FEM: Female population
AGE18PLUS_TOT: Total population age 18 years and over
AGE65PLUS_TOT: Total population age 65 years and over

```{r}
df_proc <- list()

pop_vars <- c("popestimate", "popest_fem", "age18plus_tot", "age65plus_tot")

year_code <- switch(
  as.character(POPEST_YEAR), 
  `2023` = 5, # 7/1/2023 population estimate
  `2024` = 6  # 7/1/2024 population estimate
)

df_proc$county <- df_raw %>% 
  filter(year == year_code) %>%  
  select(sumlev, state, county, all_of(pop_vars))

df_proc$state <- df_proc$county %>%
  summarize(
    across(all_of(pop_vars), \(x) sum(x, na.rm = TRUE)),
    .by = "state"
  ) %>%
  mutate(sumlev = "040", county = "000")

df_proc$us <- df_proc$state %>%
  summarize(
    across(all_of(pop_vars), \(x) sum(x, na.rm = TRUE))
  ) %>%
  mutate(sumlev = "010", state = "00", county = "000") 

df_proc <- df_proc %>%
  bind_rows() %>%
  arrange(state, county) %>%
  # construct partial UCGID (GEOID) to work with standardize_fips()
  mutate(ucgid = paste0(sumlev, "------", state, county), .before = 1) %>%
  helpers$standardize_fips() %>%
  select(statecode, countycode, all_of(pop_vars)) %>%
  helpers$add_flag_CT(old = "U", new = "A")

skimr::skim(df_proc)

```

# Calculate measures

## v051 - Population

```{r}
vname <- "v051"
measures[[vname]] <- df_proc %>%
  mutate(rawvalue = popestimate) %>%
  select(statecode, countycode, rawvalue, flag_CT) %>%
  helpers$add_col_prefix(vname)

skimr::skim(measures[[vname]])

if (SAVE_MEASURES && CHRR_YEAR == 2026) {
  save_to <- str_glue(opath$r2026_measure_)
  write_csv(measures[[vname]], save_to, na = "")
  logger::log_info("{vname} saved to {save_to}")
}

if (CHRR_YEAR == 2025) {
  x0 <- str_glue(ipath$r2025_measure_) %>%
    read_csv(show_col_types = FALSE) %>%
    arrange(statecode, countycode)
  x1 <- measures[[vname]]
  waldo::compare(x0, x1, tolerance = 1e-4, list_as_map = TRUE)
}

```


## v052 - % Below 18 Years of Age

```{r}
vname <- "v052"
measures[[vname]] <- df_proc %>%
  mutate(
    numerator = popestimate - age18plus_tot,
    denominator = popestimate,
    rawvalue = numerator / denominator
    ) %>%
  select(statecode, countycode, rawvalue, numerator, denominator, flag_CT) %>%
  helpers$add_col_prefix(vname)

skimr::skim(measures[[vname]])

if (SAVE_MEASURES && CHRR_YEAR == 2026) {
  save_to <- str_glue(opath$r2026_measure_)
  write_csv(measures[[vname]], save_to, na = "")
  logger::log_info("{vname} saved to {save_to}")
}

if (CHRR_YEAR == 2025) {
  x0 <- str_glue(ipath$r2025_measure_) %>%
    read_csv(show_col_types = FALSE) %>%
    arrange(statecode, countycode)
  x1 <- measures[[vname]]
  waldo::compare(x0, x1, tolerance = 1e-4, list_as_map = TRUE)
}

```


## v053 - % 65 and Older

```{r}
vname <- "v053"
measures[[vname]] <- df_proc %>%
  mutate(
    numerator = age65plus_tot,
    denominator = popestimate,
    rawvalue = numerator / denominator
    ) %>%
  select(statecode, countycode, rawvalue, numerator, denominator, flag_CT) %>%
  helpers$add_col_prefix(vname)


skimr::skim(measures[[vname]])

if (SAVE_MEASURES && CHRR_YEAR == 2026) {
  save_to <- str_glue(opath$r2026_measure_)
  write_csv(measures[[vname]], save_to, na = "")
  logger::log_info("{vname} saved to {save_to}")
}

if (CHRR_YEAR == 2025) {
  x0 <- str_glue(ipath$r2025_measure_) %>%
    read_csv(show_col_types = FALSE) %>%
    arrange(statecode, countycode)
  x1 <- measures[[vname]]
  waldo::compare(x0, x1, tolerance = 1e-4, list_as_map = TRUE)
}

```


## v057 - % Female

```{r}
vname <- "v057"
measures[[vname]] <- df_proc %>%
  mutate(
    numerator = popest_fem,
    denominator = popestimate,
    rawvalue = numerator / denominator
    ) %>%
  select(statecode, countycode, rawvalue, numerator, denominator, flag_CT) %>%
  helpers$add_col_prefix(vname)


skimr::skim(measures[[vname]])

if (SAVE_MEASURES && CHRR_YEAR == 2026) {
  save_to <- str_glue(opath$r2026_measure_)
  write_csv(measures[[vname]], save_to, na = "")
  logger::log_info("{vname} saved to {save_to}")
}

if (CHRR_YEAR == 2025) {
  x0 <- str_glue(ipath$r2025_measure_) %>%
    read_csv(show_col_types = FALSE) %>%
    arrange(statecode, countycode)
  x1 <- measures[[vname]]
  waldo::compare(x0, x1, tolerance = 1e-4, list_as_map = TRUE)
}

```


