---
title: "v054, v055, v056, v080, v081, v126 - R2026"
format:
  html:
    author: Anton Babkin
    date: today
    toc: true
    toc-depth: 2
    embed-resources: true
    code-fold: true
    code-overflow: wrap
    df-print: kable
---


# About

This notebook calculates the following demographic variables.

- v054, % Non-Hispanic Black
- v055, % American Indian & Alaska Native
- v056, % Hispanic
- v080, % Native Hawaiian/Other Pacific Islander
- v081, % Asian
- v126, % Non-Hispanic White

# Import packages

```{r}
library(tidyverse)
library(haven)

source("calculations/helpers.R", local = (helpers <- new.env()))
helpers$logger_config(stderr = TRUE, stdout = FALSE)

# set CHRR_YEAR to 2026 for new r2026 calculations, to 2025 for comparison against old r2025 measures
CHRR_YEAR <- 2026
POPEST_YEAR <- CHRR_YEAR - 2

# set to FALSE to run calculations without saving measures to disk
SAVE_MEASURES <- TRUE
# upload raw and processed files to S3
SAVE_TO_S3 <- FALSE
# read raw files from S3
READ_FROM_S3 <- TRUE

# calculated measures dataframes
measures <- list()

# input paths
ipath <- list(
  alldata_ = "https://www2.census.gov/programs-surveys/popest/datasets/2020-{POPEST_YEAR}/counties/asrh/cc-est{POPEST_YEAR}-alldata.csv",
  r2025_measure_ = "measure_datasets/{vname}_r2025.csv"
)

# output paths
opath <- list(
  alldata_ = "raw_data/USCB/cc-est{POPEST_YEAR}-alldata.csv",
  r2026_measure_ = "measure_datasets/{vname}_r2026.csv"
)


```



# Prepare data

- Website to downloaded data

`cc-est2023-all.csv`

https://www2.census.gov/programs-surveys/popest/datasets/2020-2023

- Online documentation

https://www2.census.gov/programs-surveys/popest/technical-documentation/file-layouts/2020-2023/CC-EST2023-ALLDATA.pdf

## import Census pop data

data: https://www2.census.gov/programs-surveys/popest/datasets/2020-2023/counties/asrh/cc-est2023-alldata.csv

```{r}
raw_file <- helpers$get_file(
  path = str_glue(opath$alldata_), 
  url = str_glue(ipath$alldata_),
  s3_read = READ_FROM_S3,
  s3_write = SAVE_TO_S3
)

df_raw <- read_csv(raw_file) %>%
  rename_with(tolower)

df_raw %>%
  head()
```


## process data

STATE: State FIPS code
COUNTY: County FIPS code
YEAR: Year
AGEGRP: Age Group
TOT_POP: Total population
NHBA_MALE: Not Hispanic, Black or African American alone male population
NHBA_FEMALE: Not Hispanic, Black or African American alone female population
NHWA_MALE: Not Hispanic, White alone male population
NHWA_FEMALE: Not Hispanic, White alone female population
IA_MALE: American Indian and Alaska Native alone male population 
IA_FEMALE: American Indian and Alaska Native alone female population
AA_MALE: Asian alone male population
AA_FEMALE: Asian alone female population
NA_MALE: Native Hawaiian and Other Pacific Islander alone male population
NA_FEMALE: Native Hawaiian and Other Pacific Islander alone female population
H_MALE: Hispanic male population
H_FEMALE: Hispanic female population


```{r}
df_proc <- list()

pop_vars <- c(
  "tot_pop", 
  "nhba_male", "nhba_female", "nhwa_male", "nhwa_female",
  "ia_male", "ia_female", "aa_male", "aa_female",
  "na_male", "na_female", "h_male", "h_female")

year_code <- switch(
  as.character(POPEST_YEAR), 
  `2023` = 5, # 7/1/2023 population estimate
  `2024` = 6  # 7/1/2024 population estimate
)

df_proc$county <- df_raw %>% 
  # agegrp == 0: total
  filter(year == year_code, agegrp == 0) %>%  
  select(sumlev, state, county, all_of(pop_vars))

df_proc$state <- df_proc$county %>%
  summarize(
    across(all_of(pop_vars), \(x) sum(x, na.rm = TRUE)),
    .by = "state"
  ) %>%
  mutate(sumlev = "040", county = "000")

df_proc$us <- df_proc$state %>%
  summarize(
    across(all_of(pop_vars), \(x) sum(x, na.rm = TRUE))
  ) %>%
  mutate(sumlev = "010", state = "00", county = "000")

df_proc <- df_proc %>%
  bind_rows() %>%
  arrange(state, county) %>%
  # construct partial UCGID (GEOID) to work with standardize_fips()
  mutate(ucgid = paste0(sumlev, "------", state, county), .before = 1) %>%
  helpers$standardize_fips() %>%
  select(statecode, countycode, all_of(pop_vars)) %>%
  helpers$add_flag_CT(old = "U", new = "A")

skimr::skim(df_proc)

```


# Calculate measures

## v054 - % Non-Hispanic Black

```{r}
vname <- "v054"
measures[[vname]] <- df_proc %>%
  mutate(
    numerator = nhba_male + nhba_female,
    denominator = tot_pop,
    rawvalue = numerator / denominator
    ) %>%
  select(statecode, countycode, rawvalue, numerator, denominator, flag_CT) %>%
  helpers$add_col_prefix(vname)

skimr::skim(measures[[vname]])

if (SAVE_MEASURES && CHRR_YEAR == 2026) {
  save_to <- str_glue(opath$r2026_measure_)
  helpers$save_to_file(measures[[vname]], save_to)
  if (SAVE_TO_S3) helpers$save_to_s3(save_to)
}

if (CHRR_YEAR == 2025) {
  x0 <- str_glue(ipath$r2025_measure_) %>%
    read_csv(show_col_types = FALSE) %>%
    arrange(statecode, countycode)
  x1 <- measures[[vname]]
  waldo::compare(x0, x1, tolerance = 1e-4, list_as_map = TRUE)
}

```


## v055 - % American Indian & Alaska Native

```{r}
vname <- "v055"
measures[[vname]] <- df_proc %>%
  mutate(
    numerator = ia_male + ia_female,
    denominator = tot_pop,
    rawvalue = numerator / denominator
    ) %>%
  select(statecode, countycode, rawvalue, numerator, denominator, flag_CT) %>%
  helpers$add_col_prefix(vname)

skimr::skim(measures[[vname]])

if (SAVE_MEASURES && CHRR_YEAR == 2026) {
  save_to <- str_glue(opath$r2026_measure_)
  helpers$save_to_file(measures[[vname]], save_to)
  if (SAVE_TO_S3) helpers$save_to_s3(save_to)
}

if (CHRR_YEAR == 2025) {
  x0 <- str_glue(ipath$r2025_measure_) %>%
    read_csv(show_col_types = FALSE) %>%
    arrange(statecode, countycode)
  x1 <- measures[[vname]]
  waldo::compare(x0, x1, tolerance = 1e-4, list_as_map = TRUE)
}

```


## v056 - % Hispanic

```{r}
vname <- "v056"
measures[[vname]] <- df_proc %>%
  mutate(
    numerator = h_male + h_female,
    denominator = tot_pop,
    rawvalue = numerator / denominator
    ) %>%
  select(statecode, countycode, rawvalue, numerator, denominator, flag_CT) %>%
  helpers$add_col_prefix(vname)

skimr::skim(measures[[vname]])

if (SAVE_MEASURES && CHRR_YEAR == 2026) {
  save_to <- str_glue(opath$r2026_measure_)
  helpers$save_to_file(measures[[vname]], save_to)
  if (SAVE_TO_S3) helpers$save_to_s3(save_to)
}

if (CHRR_YEAR == 2025) {
  x0 <- str_glue(ipath$r2025_measure_) %>%
    read_csv(show_col_types = FALSE) %>%
    arrange(statecode, countycode)
  x1 <- measures[[vname]]
  waldo::compare(x0, x1, tolerance = 1e-4, list_as_map = TRUE)
}

```


## v080 - % Native Hawaiian/Other Pacific Islander

```{r}
vname <- "v080"
measures[[vname]] <- df_proc %>%
  mutate(
    numerator = na_male + na_female,
    denominator = tot_pop,
    rawvalue = numerator / denominator
    ) %>%
  select(statecode, countycode, rawvalue, numerator, denominator, flag_CT) %>%
  helpers$add_col_prefix(vname)

skimr::skim(measures[[vname]])

if (SAVE_MEASURES && CHRR_YEAR == 2026) {
  save_to <- str_glue(opath$r2026_measure_)
  helpers$save_to_file(measures[[vname]], save_to)
  if (SAVE_TO_S3) helpers$save_to_s3(save_to)
}

if (CHRR_YEAR == 2025) {
  x0 <- str_glue(ipath$r2025_measure_) %>%
    read_csv(show_col_types = FALSE) %>%
    arrange(statecode, countycode)
  x1 <- measures[[vname]]
  waldo::compare(x0, x1, tolerance = 1e-4, list_as_map = TRUE)
}

```


## v081 - % Asian

```{r}
vname <- "v081"
measures[[vname]] <- df_proc %>%
  mutate(
    numerator = aa_male + aa_female,
    denominator = tot_pop,
    rawvalue = numerator / denominator
    ) %>%
  select(statecode, countycode, rawvalue, numerator, denominator, flag_CT) %>%
  helpers$add_col_prefix(vname)

skimr::skim(measures[[vname]])

if (SAVE_MEASURES && CHRR_YEAR == 2026) {
  save_to <- str_glue(opath$r2026_measure_)
  helpers$save_to_file(measures[[vname]], save_to)
  if (SAVE_TO_S3) helpers$save_to_s3(save_to)
}

if (CHRR_YEAR == 2025) {
  x0 <- str_glue(ipath$r2025_measure_) %>%
    read_csv(show_col_types = FALSE) %>%
    arrange(statecode, countycode)
  x1 <- measures[[vname]]
  waldo::compare(x0, x1, tolerance = 1e-4, list_as_map = TRUE)
}

```



## v126 - % Non-Hispanic White

```{r}
vname <- "v126"
measures[[vname]] <- df_proc %>%
  mutate(
    numerator = nhwa_male + nhwa_female,
    denominator = tot_pop,
    rawvalue = numerator / denominator
    ) %>%
  select(statecode, countycode, rawvalue, numerator, denominator, flag_CT) %>%
  helpers$add_col_prefix(vname)

skimr::skim(measures[[vname]])

if (SAVE_MEASURES && CHRR_YEAR == 2026) {
  save_to <- str_glue(opath$r2026_measure_)
  helpers$save_to_file(measures[[vname]], save_to)
  if (SAVE_TO_S3) helpers$save_to_s3(save_to)
}

if (CHRR_YEAR == 2025) {
  x0 <- str_glue(ipath$r2025_measure_) %>%
    read_csv(show_col_types = FALSE) %>%
    arrange(statecode, countycode)
  x1 <- measures[[vname]]
  waldo::compare(x0, x1, tolerance = 1e-4, list_as_map = TRUE)
}

```


