---
title: "v127 Premature Age-adjusted mortality (subgroups) - Rankings 2026"
author: "GL; updated by how for 2026"
date: "2025-08-02"
output: html_document
---

# Description: 

Number of deaths among residents under age 75 per 100,000 population (age-adjusted).

# Data Source: 
National Center for Health Statistics - Natality and Mortality Files 2021-2023 (requested); 

Census Population Estimates 2021-2023 from: https://www2.census.gov/programs-surveys/popest/datasets
We use births from NCHS for the infant population 

## Subgroups

- non-Hispanic, White only, 
- non-Hispanic, Black only, 
- non-Hispanic, AIAN only, 
- non-Hispanic, Asian only, 
- non-Hispanic, NHOPI only, 
- non-Hispanic, Two or More Races
- Hispanic

# import packages

```{r}
library(tidyverse)
library(haven)
```

# set up directory

```{r}
# set up directory that holds data needed for calculations
# NCHS mortality data, not allowed to share in GitHub
nchs_mort_dir <- "D:/CHRR/mortality raw"

# NCHS natality data, not allowed to share in GitHub
nchs_nata_dir <- "D:/CHRR/natality raw"

# Census county population estimates data
census_pop_dir <- "D:/CHRR/pop raw"
```

# standard county and state fips

- The final dataset will need to be aligned with standard FIPS codes and to include all counties and states.
```{r }
fips <-  bind_rows(
  read_sas("../inputs/county_fips_with_ct_old.sas7bdat"),
  read_sas( "../inputs/state_fips.sas7bdat")) %>% 
  arrange(statecode, countycode) %>% 
  glimpse()

```

# helpers

## functions used commonly by NCHS measures

- some commonly used functions are in a separate file, `nchs_measure_helpers.R`. 
```{r}
source("../scripts/nchs_measure_helpers.R")
```


## age categories for v127 calculations

```{r}
tibble::tribble(
  ~age_cat, ~`age`, 
   0L,        "0",       
   1L,      "1-4",       
   2L,      "5-14",       
   3L,    "15-24",        
   4L,    "25-34",       
   5L,    "35-44",       
   6L,    "45-54",       
   7L,    "55-64",       
   8L,    "65-74"       
  ) 
```

## subgroups: 7 total 

```{r}
tibble::tribble(
  ~subgroup, ~description, 
      
   1L,    "Non-Hispanic, White",       
   2L,    "Non-Hispanic, Black",       
   3L,    "Non-Hispanic, AIAN",       
   4L,    "Non-Hispanic, Asian",       
   5L,    "Non-Hispanic, NHOPI",       
   6L,    "Non-Hispanic, Two or more races (TOM)",   
  
   8L,    "Hispanic",  
  ) 
```

# pop data

## NCHS birth data: age_cat = 0, by subgroups

-   `nchs_births_2017_to_2023.csv`: prepared using NCHS natality data from 2017-2023

-   6 columns

    *   `race`: 1-White; 2-Black; 3-AIAN; 4-Asian; 5-NHOPI; 6-TOM
    
    *   `hisp`: 0-Not Hispanic; 1-Hispanic; 99-Unknown

```{r}
yrs <- c(2021:2023)

nchs_births <- read_csv(paste0(nchs_nata_dir, "/nchs_births_2017_2023.csv"),
                        show_col_types = FALSE) %>% 
  filter(year %in% yrs) %>% 
  glimpse()

infant_pop <- nchs_births %>% 
  # exclude Hispanic unknown
  filter(hisp != 99) %>% 
  # set race/eth groups as 6 + 1
  mutate(race = case_when(
    hisp == 0 & race == 1 ~ 1, # non-Hispanic, "White",
    hisp == 0 & race == 2 ~ 2, # non-Hispanic, "Black",
    hisp == 0 & race == 3 ~ 3, # non-Hispanic, "AIAN", 
    hisp == 0 & race == 4 ~ 4, # non-Hispanic, "Asian",
    hisp == 0 & race == 5 ~ 5, # non-Hispanic, "NHOPI", 
    hisp == 0 & race == 6 ~ 6, # non-Hispanic, "Two or More Races"
    hisp == 1  ~  8,           # Hispanic
    TRUE ~ NA
  )) %>% 
  group_by(statecode, countycode, race) %>% 
  summarise(pop = sum(births, na.rm = TRUE), age_cat = 0, .groups = "drop") %>% 
  glimpse()
```

## Census pop: age_cat 1-8, by subgroups

### function to get pop by by race, age cat 1-8

```{r}
get_pop_by_sub61_agecat_v127 <- function(census_ct_estimates, year_num = 13, long = TRUE) {
  
  print(paste0("YEAR = ", year_num))
  
  cc_agegrp_sg_r <- census_ct_estimates %>%  
    filter(year == year_num, agegrp != 0) %>% 
    # select columns for single races
    select(statecode = state, countycode = county, agegrp,
           # Non-Hispanic: 6
           starts_with("nhwa_"), # Not Hispanic, White alone
           starts_with("nhba_"), # Not Hispanic, Black or African American alone
           starts_with("nhia_"), # Not Hispanic, American Indian and Alaska Native alone
           starts_with("nhaa_"), # Not Hispanic, Asian alone
           starts_with("nhna_"), # Not Hispanic, Native Hawaiian and Other Pacific Islander alone
           starts_with("nhtom_"), # NHTOM_MALE	Not Hispanic, Two or More Races male population
           # Hispanic: 1
           starts_with("h_"), # Hispanic
    ) %>%  
    mutate(
      # 6: Non-Hispanic groups
      nh_white = nhwa_male  + nhwa_female,
      nh_black = nhba_male  + nhba_female,
      nh_aian  = nhia_male  + nhia_female,
      nh_asian = nhaa_male  + nhaa_female,
      nh_nhopi = nhna_male  + nhna_female,
      nh_tom   = nhtom_male + nhtom_female,
      # 1: Hispanic 
      hispanic = h_male  + h_female
    ) %>% 
    select(-c(nhwa_male:h_female)) %>% 
    # fix counties
    mutate(countycode = case_when(
      statecode=='46' & countycode =='113' ~ '102',
      statecode=='51' & countycode =='515' ~ '019',
      statecode=='02' & countycode =='270' ~ '158',
      statecode=='02' & countycode =='201' ~ '198',
      TRUE ~ countycode
    )) 

   # county pop
  # 0-74: age cat 1-8; will fix age_cat 1
  ct_pop_agecat_1_8 <- cc_agegrp_sg_r %>%  
    filter(agegrp %in% c(1:15)) %>% 
    mutate(age_cat = case_when(
      agegrp %in% c(1) ~ 1, # 0-4; will be fixed using NCHS birth data
      agegrp %in% c(2, 3) ~ 2, # 5-9, 10-14    
      agegrp %in% c(4, 5) ~ 3, # 15-19, 19-24  
      agegrp %in% c(6, 7) ~ 4, # 25-29, 30-34
      agegrp %in% c(8, 9) ~ 5, # 35-39, 40-44
      agegrp %in% c(10, 11) ~ 6, # 45-49, 50-54
      agegrp %in% c(12, 13) ~ 7, # 55-59, 60-64
      agegrp %in% c(14, 15) ~ 8, # 65-69, 70-74
      TRUE ~ NA
    )) %>% 
    select(-agegrp) %>% 
    group_by(statecode, countycode, age_cat) %>% 
    summarise(across(c(nh_white:hispanic), ~sum(., na.rm = TRUE)), .groups = "drop")
  
  # state pop
  st_pop_agecat_1_8 <- ct_pop_agecat_1_8 %>% 
    group_by(statecode, age_cat) %>% 
    summarise(across(c(nh_white:hispanic), ~sum(., na.rm = TRUE)), .groups = "drop") %>% 
    mutate(countycode = "000", .after = 1)
  
  # us pop
  us_pop_agecat_1_8 <- ct_pop_agecat_1_8 %>% 
    group_by(age_cat) %>% 
    summarise(across(c(nh_white:hispanic), ~sum(., na.rm = TRUE)), .groups = "drop") %>% 
    mutate(statecode = "00", countycode = "000", .before = 1)
  
  ## county + state + us data
  pop_all_agecat_1_8 <- bind_rows(
    us_pop_agecat_1_8, 
    st_pop_agecat_1_8, 
    ct_pop_agecat_1_8) %>% 
    arrange(statecode, countycode, age_cat)
  
  
  # long form or wide form
  if(long){
    
    pop_all_agecat_1_8_long <- pop_all_agecat_1_8 %>% 
      pivot_longer(cols = c(nh_white:hispanic), names_to = "race", values_to = "pop") %>% 
      mutate(race = case_when(
        race == "nh_white" ~ 1,
        race == "nh_black" ~ 2,
        race == "nh_aian"  ~ 3,
        race == "nh_asian" ~ 4,
        race == "nh_nhopi" ~ 5,
        race == "nh_tom"   ~ 6,
        
        race == "hispanic"  ~ 8,
        TRUE ~ NA_real_))%>% 
      arrange(statecode, countycode, race, age_cat)
    
    return(pop_all_agecat_1_8_long)
    
  }
  else{
    return(pop_all_agecat_1_8)
  }
  
}
```

### function to process Census pop data for v127

```{r}
process_pop_v127 <- function(year) {
  pop <- read_census_pop(
    file_name = get_census_csv_file_info(year)$file_name
  ) %>%
    get_pop_by_sub61_agecat_v127(
      .,
      year_num = get_census_csv_file_info(year)$year_num
    )

  return(pop)
}
```


## pop: age_cat 1-8 for 2021, 2022, 2023 by subgroups

```{r}
pop_agecat_1_8 <- bind_rows(  
  process_pop_v127(2021), 
  process_pop_v127(2022)%>% fix_CT_pop(process_pop_v127(2021)),
  process_pop_v127(2023) %>% fix_CT_pop(process_pop_v127(2021)),
  ) %>%  
  group_by(statecode, countycode, age_cat, race) %>%  
  summarise(pop = round(sum(pop, na.rm = TRUE), digits = 0), .groups = "drop") %>% 
  # remove 02261
  filter(!(statecode == "02" & countycode =="261"  ) )

pop_agecat_1_8
```


## pop_all: add age_cat 0 (infant), fix age_cat 1

```{r}
# fix age_cat 1: subtract infant pop from NCHS births data
pop_agecat_1_8 %>% 
  filter(age_cat==1) %>% 
  bind_rows(infant_pop) %>% 
  arrange(statecode, countycode, age_cat, race) %>% 
  pivot_wider(names_from = age_cat, values_from = pop, names_glue = "pop_{age_cat}") %>% 
  mutate(pop = if_else(is.na(pop_0), pop_1, pop_1 - pop_0)) %>% 
  select(-c(pop_0, pop_1)) %>% 
  mutate(age_cat = 1)

```



```{r}
# age_cat = 0, for all counties, set missing as 0
fips %>% 
  select(1:2) %>% 
  mutate(age_cat = 0) %>% 
  left_join(infant_pop, by = c('statecode', 'countycode', 'age_cat')) %>% 
  mutate(pop = if_else(is.na(pop), 0, pop))

```



```{r}
pop_all <- 
  bind_rows(
    # age_cat = 2 to 18
    pop_agecat_1_8 %>% filter(age_cat!=1),
    
    # age_cat = 0
    c(1:6, 8) %>% 
      map_df(~mutate(fips %>% select(1:2), race = .)) %>% 
      mutate(age_cat = 0) %>% 
      left_join(infant_pop, by = c('statecode', 'countycode', 'race', 'age_cat')) %>% 
      mutate(pop = if_else(is.na(pop), 0, pop)),
    
    # fixed age_cat = 1
    pop_agecat_1_8 %>% 
      filter(age_cat==1) %>% 
      bind_rows(infant_pop) %>% 
      pivot_wider(names_from = age_cat, values_from = pop, names_glue = "pop_{age_cat}") %>% 
      mutate(pop = if_else(is.na(pop_0), pop_1, pop_1 - pop_0)) %>% 
      select(-c(pop_0, pop_1)) %>% 
      mutate(age_cat = 1)
    ) %>% 
  arrange(statecode, countycode, age_cat, race) %>% 
  # set negative pop as 0; example: 15005
  mutate(pop = if_else(pop<0, 0, pop)) %>%
  # remove AK 02261
  filter(!(statecode == "02" & countycode == "261") ) %>% 
  glimpse()

pop_all  
```


# mort data

-   Note: for convenience,  all mortality data files were named in the format as `MULTYEARUS.AllCnty.txt`, e.g., `MULT2019US.AllCnty.txt`

## header

```{r}
mort_header <- tribble(
  ~fld_name, ~start,  ~end,
  "state_of_residence", 29,  30,
  "county_of_residence", 35,  37,
  "detail_age", 70, 73,
  # "icd_code", 146, 149,
  "race", 445, 446,
  "race_recode_5", 450, 450,
  "hispanic_origin", 484, 486,
  "hispanic_origin_race_recode", 488, 488,
  "race_recode_40", 489, 490
)
```

## function to get age category for mort data
```{r}
get_age_cat_v127 <- function(age) {
  case_when(
    age == 0 ~ 0,
    age <= 4 ~ 1,
    age <= 14 ~ 2,
    age <= 24 ~ 3,
    age <= 34 ~ 4,
    age <= 44 ~ 5,
    age <= 54 ~ 6,
    age <= 64 ~ 7,
    age <= 74 ~ 8,
    TRUE ~ 99
  )
}
```

## function to process single year mort data

```{r}
get_mort_by_sub61_agecat_v127 <- function(nchs_mort, df_fips){
  mort_1 <- nchs_mort %>% 
      # include only 50 states + DC
    filter(state_of_residence %in% c(state.abb, "DC")) %>% 
    rename(state = state_of_residence, countycode = county_of_residence) %>% 
    # fix county FIPS codes
    mutate(countycode = case_when(
      state =='SD' & countycode =='113' ~'102',
      state =='AK' & countycode =='270' ~'158',
      state =='VA' & countycode =='515' ~'019',
      state =='AK' & countycode =='063' ~'261',
      state =='AK' & countycode =='066' ~'261',
      TRUE ~ countycode
      )
    )
  
  mort_2 <- mort_1 %>% 
    # remove age not stated
    filter(detail_age!= 9999, detail_age!=1999) %>% 
    mutate(age_death = case_when(
      detail_age >1999 ~ 0,
      detail_age>1000 & detail_age<2000 ~ detail_age-1000,
      TRUE ~ NA_real_
    )) %>% 
    filter(!is.na(age_death)) %>% 
    select(-c("detail_age")) %>% 
    mutate(age_death <75) %>% 
    mutate(age_cat = get_age_cat_v127(age_death)) %>% 
    # race
    select(-race) %>% 
    mutate(race_recode_40 = as.numeric(race_recode_40)) %>% 
    mutate(
    race = case_when(
      # non-Hispanic
    	hispanic_origin<200 & race_recode_40==1 ~ 1, # nh_white
    	hispanic_origin<200 & race_recode_40==2 ~ 2, # nh_black
    	hispanic_origin<200 & race_recode_40==3 ~ 3, # nh_aian
    	hispanic_origin<200 & race_recode_40 %in% c(4:10)  ~ 4, # nh_Asian
    	hispanic_origin<200 & race_recode_40 %in% c(11:14) ~ 5, # nh_NHOPI
    	hispanic_origin<200 & race_recode_40 %in% c(15:40) ~ 6, # nh_two or more races
    	
      # Hispanic
  	  hispanic_origin<300  ~ 8, # Hispanic

    	TRUE ~ NA_real_
    )
  )
  
  # county
  mort_ct <- mort_2 %>% 
    group_by(state, countycode, age_cat, race) %>% 
    summarise(deaths = n(), .groups = "drop")
  
  # state
  mort_st <- mort_2 %>% 
    group_by(state, age_cat, race) %>% 
    summarise(countycode = "000", deaths = n(), .groups = "drop") 
  
  # us
  mort_us <- mort_2 %>% 
    group_by(age_cat, race) %>% 
    summarise(state = "US", countycode = "000", deaths = n(), .groups = "drop")
  
  # county + state + us
  mort_all <- bind_rows(mort_ct, mort_st, mort_us) %>% 
  arrange(state, countycode) %>% 
  # add statecode
  left_join(df_fips %>% select(state, statecode) %>% 
              distinct(),
            by = c("state")) %>% 
  ungroup() %>% 
  select(-state) %>% 
  select(statecode, everything()) %>% 
  arrange(statecode, countycode, race, age_cat) %>% 
  filter(age_cat != 99) 
  
  return(mort_all)
  
}
```

## function to process mortality data for v127 subgroups

-   a helper that combines reading and processing one mortality dataset

```{r}
process_mort_v127 <- function(year){
  
  print(get_mort_file_name(year))
  
  mort <- read_mort_data(
    file_path = file.path(nchs_mort_dir, get_mort_file_name(year) ), 
    header = mort_header,
    n_max = Inf) %>% 
    get_mort_by_sub61_agecat_v127(df_fips=fips)
  
  return(mort)
}
```





## mort 3 years
```{r}
mort_all <- bind_rows(
  process_mort_v127(2021),
  process_mort_v127(2022),
  process_mort_v127(2023)
  ) %>% 
  group_by(statecode, countycode, age_cat, race) %>% 
  summarise(deaths = sum(deaths), .groups = "drop") %>% 
  glimpse()

```



## merge mort and pop data

```{r}
merged <- pop_all %>% 
  left_join(mort_all %>% ungroup(), 
            by = c("statecode", "countycode", "age_cat", "race")) %>% 
  # replace NA with 0 for deaths
  mutate(deaths = if_else(is.na(deaths), 0, deaths)) %>% 
  glimpse()
  
```

## function to assign age weight numerator based on age category

	if age_cat=0 then wtn=3794901;
	if age_cat=1 then wtn=15191619; /*1-4 years*/
	if age_cat=2 then wtn=39976619; /*5-14 years*/
	if age_cat=3 then wtn=38076743; /*15-24 years*/
	if age_cat=4 then wtn=37233437; /*25-34 years*/
	if age_cat=5 then wtn=44659185; /*35-44 years*/
	if age_cat=6 then wtn=37030152; /*45-54 years*/
	if age_cat=7 then wtn=23961506; /*55-64 years*/
	if age_cat=8 then wtn=18135514; /*65-74 years*/
	wtd=258059676;
	
```{r}
# function to assign age weight numerator based on age category -----------
get_age_wtn_v127 <- function(age_cat){
  case_when(
  	age_cat==0 ~ 3794901, #/*under 1*/
  	age_cat==1 ~ 15191619,# /*1-4 years*/
  	age_cat==2 ~ 39976619,# /*5-14 years*/
  	age_cat==3 ~ 38076743,# /*15-24 years*/
  	age_cat==4 ~ 37233437,# /*25-34 years*/
  	age_cat==5 ~ 44659185,# /*35-44 years*/
  	age_cat==6 ~ 37030152,# /*45-54 years*/
  	age_cat==7 ~ 23961506,# /*55-64 years*/
  	age_cat==8 ~ 18135514,# /*65-74 years*/
    TRUE ~ NA_real_
  )
}
```

## add weight

```{r}
wtd <- 258059676 # /*population under 75; Census 2000 model*/

merged_2 <- merged %>% 
  mutate(wtn = get_age_wtn_v127(age_cat)) %>% 
  mutate(wt = wtn/wtd,
         age_adj=deaths/pop*wtn/wtd*100000,
         no_age_adj=deaths/pop*100000,
         wk_var = (wt**2)*((no_age_adj**2)/(deaths)))

merged_3 <- merged_2 %>% 
  group_by(statecode, countycode,race) %>% 
  summarise(numerator = sum(deaths, na.rm = TRUE),
            denominator = sum(pop, na.rm = TRUE),
            rawvalue = sum(age_adj, na.rm = TRUE), 
            # calculate std
            stdev = sqrt(sum(wk_var, na.rm = TRUE)),
            .groups = "drop") %>% 
  glimpse()
```



## confidence Intervals

```{r}
v127_1 <- merged_3 %>% 
  mutate(cilow = rawvalue - 1.96*stdev, 
         cihigh = rawvalue + 1.96*stdev,
         relative_standard_error = (stdev/rawvalue)) %>% 
  # select(-stdev) %>% 
  mutate(flag2 = if_else(relative_standard_error > 0.25, 
                         1, NA_real_)) %>% 
  mutate(flag2 = if_else(numerator < 20, 2, flag2)) %>% 
  # rename(v127_numerator = numerator, v127_denominator = denominator,
  #        v127_rawvalue = rawvalue, v127_cilow = cilow, v127_cihigh= cihigh) %>% 
  mutate(across(c(numerator, denominator, rawvalue), ~ as.double(.x))) %>% 
  mutate(across(c(numerator, denominator, rawvalue), ~ if_else(!is.na(flag2) & flag2 == 2, 
                                            NA_real_, .x))) %>% 
  mutate(test_cb=round((rawvalue**2)/(stdev**2), digits = 0))

v127_1
```


## calculate CI L and U

```{r}
df_ci <- get_pois_CI() %>% 
  glimpse()
```




```{r}
v127_2 <- v127_1 %>% 
  left_join(df_ci, by = c("test_cb" = "n")) %>% 
  # mutate(flag2 = if_else(v001_numerator<20, 
  #                        2, flag2)) %>%
  mutate(cilow = if_else(numerator>=20 &numerator <=99,
                         rawvalue * L, cilow),
         cihigh = if_else(numerator>=20 & numerator <=99,
                         rawvalue * U,cihigh)) %>% 
  # mutate(v001_sourceflag = NA_real_) %>% 
  # mutate(flag2 = if_else(relative_standard_error > 0.2 & flag2!=2, 
  #                        1, flag2)) %>% 
  select(-c(relative_standard_error, test_cb, L, U)) %>% 
  mutate(across(starts_with("v"), ~ if_else(!is.na(flag2) & flag2 == 2, 
                                            NA_real_, .x))) %>% 
  rename(ypll_se=stdev, flag2_cb = flag2) %>% 
  # fix se for 15005
  mutate(ypll_se = if_else(statecode=="15" & countycode =="005",
                           NA_real_, ypll_se)) %>% 
  select(-c(ypll_se, flag2_cb)) %>% 
  mutate(sourceflag = NA_real_)

v127_2
```




## get subgroup data in wide format

### function to get a subgroup

```{r}
get_race_data <- function(df, race_cat = 1, race_group = "v127_race_white"){
  v127_2 %>% 
    filter(race == race_cat) %>% 
    select(-c(race, numerator, denominator, sourceflag)) %>% 
    # replace inf with NA
  mutate(across(c('rawvalue', 'cilow', 'cihigh'), ~if_else(is.infinite(.), NA, .))) %>% 
    rename({{race_group}} := rawvalue) %>% 
    rename_with(.fn = ~paste0(race_group, "_", .x), .cols = c("cilow" , "cihigh" ))
     
}
```

### subgroup data

```{r}
white <- get_race_data(v127_2, race_cat = 1, race_group = "v127_race_white")
black <- get_race_data(v127_2, race_cat = 2, race_group = "v127_race_black")
aian  <- get_race_data(v127_2, race_cat = 3, race_group = "v127_race_aian")
asian <- get_race_data(v127_2, race_cat = 4, race_group = "v127_race_asian")
nhopi <- get_race_data(v127_2, race_cat = 5, race_group = "v127_race_nhopi")
tom   <- get_race_data(v127_2, race_cat = 6, race_group = "v127_race_tom")

hispanic <- get_race_data(v127_2, race_cat = 8, race_group = "v127_race_hispanic")

```

### bind columns

```{r}
v127_other <- purrr::reduce(
  list(fips %>%  
         select(1:2), 
       white, black, aian, asian, nhopi, tom,
       hispanic
       ), 
  dplyr::left_join, 
  by = c("statecode", "countycode")) 

v127_other
```

## add flag_CT

```{r}
v127_other_2 <- v127_other %>%
  add_flag_CT(vnum="v127_race", old = "U", new = "A")

```

# compare initial and duplicated calculations

```{r}
compare_subgroup("v127", v127_other_2, 2026)
```

```{r}
# one row is set to missing in GL's data that is not NA in my data - looks like it should be suppressed. 
problem_rows <- which(is.na(initial$v127_race_black) & !is.na(v127_other_2$v127_race_black))
v127_other_2[problem_rows, ]

# Set the specified columns to NA in v127_other_2
v127_other_2$v127_race_black[problem_rows] <- NA
v127_other_2$v127_race_black_cilow[problem_rows] <- NA
v127_other_2$v127_race_black_cihigh[problem_rows] <- NA
```


# save data
```{r}
save_data(v127_other_2, "v127_otherdata")
```
