---
title: "v129 Infant mortality - Ranking 2026"
output: html_document

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

# about

## Basic information

- Description
Number of infant deaths (within 1 year) per 1,000 live births.

- What is the numerator?
The numerator is the cumulative number of deaths occurring before one year of age.

- What is the denominator?
The denominator is the total number of live births.

# import packages

```{r}
library(tidyverse)
library(haven)
```

# set up directory

```{r}
# set up directory that holds data needed for calculations
# NCHS mortality data, not allowed to share in GitHub
nchs_mort_dir <- "D:/CHRR/mortality raw"

# NCHS natality data, not allowed to share in GitHub
nchs_nata_dir <- "D:/CHRR/natality raw"

# Census county population estimates data
census_pop_dir <- "D:/CHRR/pop raw"
```

# standard county and state fips

- The final dataset will need to be aligned with standard FIPS codes and to include all counties and states.
```{r }
fips <-  bind_rows(
  read_sas("../inputs/county_fips_with_ct_old.sas7bdat"),
  read_sas( "../inputs/state_fips.sas7bdat")) %>% 
  arrange(statecode, countycode) %>% 
  glimpse()

```

# helpers

## functions used commonly by NCHS measures

- some commonly used functions are in a separate file, `nchs_measure_helpers.R`. 
```{r}
source("../scripts/nchs_measure_helpers.R")
```

## function to process 1 year mort data for infants

```{r}
get_infant_mort <- function(nchs_mort, df_fips){
  mort_1 <- nchs_mort %>% 
      # include only 50 states + DC
    filter(state_of_residence %in% c(state.abb, "DC")) %>% 
    rename(state = state_of_residence, countycode = county_of_residence) %>% 
    # fix county FIPS codes
    mutate(countycode = case_when(
      state =='SD' & countycode =='113' ~'102',
      state =='AK' & countycode =='270' ~'158',
      state =='VA' & countycode =='515' ~'019',
      TRUE ~ countycode
      )
    )
  
  mort_2 <- mort_1 %>% 
    filter(detail_age!= 9999 & detail_age >1999) %>% 
    select(-c("detail_age")) 
  
  # county
  mort_ct <- mort_2 %>% 
    group_by(state, countycode) %>% 
    summarise(deaths = n(), .groups = "drop")
  
  # state
  mort_st <- mort_2 %>% 
    group_by(state) %>% 
    summarise(countycode = "000", deaths = n(), .groups = "drop") 
  
  # us
  mort_us <- mort_2 %>% 
    summarise(state = "US", countycode = "000", deaths = n())
  
  mort_all <- bind_rows(mort_ct, mort_st, mort_us) %>% 
  arrange(state, countycode) %>% 
  # add statecode
  left_join(df_fips %>% select(state, statecode) %>% 
              distinct(),
            by = c("state")) %>% 
  ungroup() %>% 
  select(-state) %>% 
  select(statecode, everything()) %>% 
  arrange(statecode, countycode) 
  
  return(mort_all)
  
}
```

## function to process mortality data for v129 subgroups

```{r}
process_mort_v129 <- function(year){
  
  print(get_mort_file_name(year))
  
  mort <- read_mort_data(
    file_path = file.path(nchs_mort_dir, get_mort_file_name(year) ), 
    header = mort_header,
    n_max = Inf) %>% 
    get_infant_mort(df_fips=fips)
  
  return(mort)
}
```

# Pop data: NCHS birth data

age_cat = 0, by race 6 + 6

   1L,   "White", non-Hispanic
   2L,   "Black", non-Hispanic
   3L,   "AIAN", non-Hispanic
   4L,   "Asian", non-Hispanic
   5L,   "NHOPI", non-Hispanic
   6L,   "Two or More Races", non-Hispanic

   11L,   "White", Hispanic
   12L,   "Black", Hispanic
   13L,   "AIAN", Hispanic
   14L,   "Asian", Hispanic
   15L,   "NHOPI", Hispanic
   16L,   "Two or More Races", Hispanic

```{r}
yrs <- c(2017:2023)

nchs_births <- read_csv(paste0(nchs_nata_dir, "/nchs_births_2017_2023.csv")) %>% 
  filter(year %in% yrs) %>% 
  glimpse()

infant_pop <- nchs_births %>% 
  group_by(statecode, countycode) %>% 
  summarise(pop = sum(births, na.rm = TRUE), .groups = "drop") %>% 
  arrange(statecode, countycode) %>% 
  # select(1:2, age_cat, race, pop) %>% 
  glimpse()

infant_pop
```

```{r}
pop_all <- infant_pop %>% 
  # fix counties
  mutate(countycode = case_when(
    statecode=='46' & countycode =='113' ~ '102',
    statecode=='51' & countycode =='515' ~ '019',
    statecode=='02' & countycode =='270' ~ '158',
    statecode=='02' & countycode =='201' ~ '198',
    TRUE ~ countycode
  )) %>% 
  right_join(
    fips %>% 
      select(statecode, countycode),
    by = c("statecode", "countycode")
  ) %>% 
  arrange(statecode, countycode) %>% 
  glimpse()
```

# mortality data

## header

```{r}
mort_header <- tribble(
  ~fld_name, ~start,  ~end,
  "state_of_residence", 29,  30,
  "county_of_residence", 35,  37,
  "detail_age", 70, 73
)
```

## mort 2017-2023: map

```{r}
mort_all <- c(2017:2023) %>% 
  map(process_mort_v129) %>% 
  list_rbind() %>% 
  group_by(statecode, countycode) %>% 
  summarise(deaths = sum(deaths), .groups = "drop") %>%
  glimpse()

```

# mortality calculation

## merge mort and pop data

```{r}
merged <- pop_all %>% 
  left_join(mort_all, 
            by = c("statecode", "countycode")) %>% 
  glimpse()
  
```

## confidence Intervals

```{r}
v129 <- merged %>%
  rename(numerator = deaths, denominator = pop) %>%
  mutate(
    rawvalue = numerator / denominator *1000,
    rse = 100 * sqrt(1/numerator),
  ) %>%
  mutate(
    std = rawvalue * rse / 100,
    cilow = rawvalue - 1.96 * std,
    cihigh = rawvalue + 1.96 * std
  ) %>%
  select(-c(rse, std)) %>% 
  glimpse()

```

## calculate CI L and U for n < 100

```{r}
v129_2 <- v129 %>% 
  left_join(get_pois_CI(), by = c("numerator" = "n")) %>% 
  mutate(cilow = if_else(numerator>=20 & numerator <=99,
                         rawvalue * L, cilow),
         cihigh = if_else(numerator>=20 & numerator <=99,
                         rawvalue * U, cihigh)) %>% 
  select(-c(L, U)) %>% 
  mutate(across(where(is.numeric), ~ if_else(numerator < 20, 
                                            NA_real_, .x))) %>% 
  mutate(sourceflag  = NA_real_) %>% 
  rename_with(.fn = ~paste0("v129_", .x), .cols = c("denominator", "numerator","rawvalue", "cilow" , "cihigh" ,"sourceflag"))

v129_2

```

# add flag_CT

- 'A' = data *available* for a CT county
- 'U' = data *unavailable* for a CT county

```{r}
v129_3 <- v129_2 %>% 
  add_flag_CT(vnum = "v129", old="A", new = "U")
```

# Compare to initial calculations 
```{r}
compare("v129", v129_3, year = 2026)
```


# save data
```{r}
save_data(v129_3, "v129")
```
