---
title: "v129 Infant mortality_otherdata (subgroups) - Ranking 2026"
output: html_document

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

# about

## Basic information

- Description
Number of infant deaths (within 1 year) per 1,000 live births.

- What is the numerator?
The numerator is the cumulative number of deaths occurring before one year of age.

- What is the denominator?
The denominator is the total number of live births.

## Subgroups

- non-Hispanic, White only, 
- non-Hispanic, Black only, 
- non-Hispanic, AIAN only, 
- non-Hispanic, Asian only, 
- non-Hispanic, NHOPI only, 
- non-Hispanic, Two or More Races
- Hispanic

# import packages

```{r}
library(tidyverse)
library(haven)
```

# set up directory

```{r}
# set up directory that holds data needed for calculations
# NCHS mortality data, not allowed to share in GitHub
nchs_mort_dir <- "D:/CHRR/mortality raw"

# NCHS natality data, not allowed to share in GitHub
nchs_nata_dir <- "D:/CHRR/natality raw"

# Census county population estimates data
census_pop_dir <- "D:/CHRR/pop raw"
```

# standard county and state fips

- The final dataset will need to be aligned with standard FIPS codes and to include all counties and states.
```{r }
fips <-  bind_rows(
  read_sas("../inputs/county_fips_with_ct_old.sas7bdat"),
  read_sas( "../inputs/state_fips.sas7bdat")) %>% 
  arrange(statecode, countycode) %>% 
  glimpse()

```

# helpers

## functions used commonly by NCHS measures

- some commonly used functions are in a separate file, `nchs_measure_helpers.R`. 
```{r}
source("../scripts/nchs_measure_helpers.R")
```

## function to process single year mort data for infants by race

```{r}
get_infant_mort_sub61 <- function(nchs_mort, df_fips){
  mort_1 <- nchs_mort %>% 
    # keep only infant deaths
    filter(detail_age!= 9999 & detail_age >1999) %>% 
    select(-c("detail_age")) %>% 
      # include only 50 states + DC
    filter(state_of_residence %in% c(state.abb, "DC")) %>% 
    rename(state = state_of_residence, countycode = county_of_residence) %>% 
    # fix county FIPS codes
    mutate(countycode = case_when(
      state =='SD' & countycode =='113' ~'102',
      state =='AK' & countycode =='270' ~'158',
      state =='VA' & countycode =='515' ~'019',
      TRUE ~ countycode
      )
    )
  
  mort_2 <- mort_1 %>% 
    # race
    mutate(race_recode_40 = as.numeric(race_recode_40)) %>% 
    mutate(
      race = case_when(
      # non-Hispanic
      hispanic_origin<200 & race_recode_40==1 ~ 1, # nh_white
    	hispanic_origin<200 & race_recode_40==2 ~ 2, # nh_black
    	hispanic_origin<200 & race_recode_40==3 ~ 3, # nh_aian
    	hispanic_origin<200 & race_recode_40 %in% c(4:10)  ~ 4, # nh_Asian
    	hispanic_origin<200 & race_recode_40 %in% c(11:14) ~ 5, # nh_NHOPI
    	hispanic_origin<200 & race_recode_40 %in% c(15:40) ~ 6, # nh_two or more races
    	
      # Hispanic
  	  hispanic_origin<300  ~ 8, # Hispanic

    	TRUE ~ NA_real_
      )
    )
  
  # county
  mort_ct <- mort_2 %>% 
    group_by(state, countycode, race) %>% 
    summarise(deaths = n(), .groups = "drop")
  
  # state
  mort_st <- mort_2 %>% 
    group_by(state, race) %>% 
    summarise(countycode = "000", deaths = n(), .groups = "drop") 
  
  # us
  mort_us <- mort_2 %>%  
    group_by(race) %>% 
    summarise(state = "US", countycode = "000", deaths = n())
  
  mort_all <- bind_rows(mort_ct, mort_st, mort_us) %>% 
  arrange(state, countycode) %>% 
  # add statecode
  left_join(df_fips %>% select(state, statecode) %>% 
              distinct(),
            by = c("state")) %>% 
  ungroup() %>% 
  select(-state) %>% 
  select(statecode, everything()) %>% 
  arrange(statecode, countycode, race) 
  
  return(mort_all)
  
}
```

## function to process mortality data for v129 subgroups

```{r}
process_mort_v129 <- function(year){
  
  print(get_mort_file_name(year))
  
  mort <- read_mort_data(
    file_path = file.path(nchs_mort_dir, get_mort_file_name(year) ), 
    header = mort_header,
    n_max = Inf) %>% 
    get_infant_mort_sub61(df_fips=fips)
  
  return(mort)
}
```

# pop data: NCHS birth data

```{r}
yrs <- c(2017:2023)

nchs_births <- read_csv(paste0(nchs_nata_dir, "/nchs_births_2017_2023.csv")) %>% 
  filter(year %in% yrs) %>% 
  glimpse()

infant_pop <- nchs_births %>% 
  # exclude Hispanic unknown
  filter(hisp != 99) %>% 
  # set race/eth groups as 6 + 1
  mutate(race = case_when(
    hisp == 0 & race == 1 ~ 1, # non-Hispanic, "White",
    hisp == 0 & race == 2 ~ 2, # non-Hispanic, "Black",
    hisp == 0 & race == 3 ~ 3, # non-Hispanic, "AIAN", 
    hisp == 0 & race == 4 ~ 4, # non-Hispanic, "Asian",
    hisp == 0 & race == 5 ~ 5, # non-Hispanic, "NHOPI", 
    hisp == 0 & race == 6 ~ 6, # non-Hispanic, "Two or More Races"
    hisp == 1  ~  8,           # Hispanic
    TRUE ~ NA
  )) %>% 
  group_by(statecode, countycode, race) %>% 
  summarise(pop = sum(births, na.rm = TRUE), age_cat = 0, .groups = "drop") %>% 
  arrange(statecode, countycode, age_cat, race) %>% 
  select(1:2, age_cat, race, pop) %>% 
  glimpse()

infant_pop
```

```{r}
infant_pop_2 <- infant_pop %>% 
  select(-age_cat) %>% 
  # fix counties
  mutate(countycode = case_when(
    statecode=='46' & countycode =='113' ~ '102',
    statecode=='51' & countycode =='515' ~ '019',
    statecode=='02' & countycode =='270' ~ '158',
    statecode=='02' & countycode =='201' ~ '198',
    TRUE ~ countycode
  ))

pop_all <- c(1:6, 8) %>% 
  map_df(~mutate(fips %>% select(1:2), race = .)) %>% 
  left_join(infant_pop_2, by = c('statecode', 'countycode', 'race')) %>% 
  mutate(pop = if_else(is.na(pop), 0, pop)) %>% 
  glimpse()
```

# mortality data

## header

```{r}
mort_header <- tribble(
  ~fld_name, ~start,  ~end,
  "state_of_residence", 29,  30,
  "county_of_residence", 35,  37,
  "detail_age", 70, 73,
  "hispanic_origin", 484, 486,
  "race_recode_40", 489, 490
)
```

## mort 7 years: 2017-2023

```{r}
mort_all <- c(2017:2023) %>% 
  map(process_mort_v129) %>% 
  list_rbind() %>% 
  group_by(statecode, countycode, race) %>%
  summarise(deaths = sum(deaths), .groups = "drop") %>%
  filter(!is.na(race)) %>%
  glimpse()

```

# mortality calculation

## merge mort and pop data

```{r}
merged <- pop_all %>% 
  left_join(mort_all, 
            by = c("statecode", "countycode", "race")) %>% 
  glimpse()
  
```

## confidence Intervals

```{r}
v129 <- merged %>%
  filter(race != 99) %>% 
  rename(numerator = deaths, denominator = pop) %>%
  mutate(
    rawvalue = numerator / denominator *1000,
    rse = 100 * sqrt(1/numerator),
  ) %>%
  mutate(
    std = rawvalue * rse / 100,
    cilow = rawvalue - 1.96 * std,
    cihigh = rawvalue + 1.96 * std
  ) %>%
  select(-c(rse, std)) %>% 
  glimpse()

```

## calculate CI L and U for n < 100

```{r}
v129_2 <- v129 %>% 
  left_join(get_pois_CI(), by = c("numerator" = "n")) %>% 
  mutate(cilow = if_else(numerator>=20 & numerator <=99,
                         rawvalue * L, cilow),
         cihigh = if_else(numerator>=20 & numerator <=99,
                         rawvalue * U, cihigh)) %>% 
  select(-c(L, U)) %>% 
  mutate(across(c(denominator:cihigh), ~ if_else(numerator < 20, 
                                            NA_real_, .x))) %>%  
  glimpse()
```


## get subgroups

```{r}
v129_other <- process_race_data_wide(v129_2, fips, "v129") %>% 
  select(-c(ends_with('denominator'), ends_with('numerator')))

```

## add flag_CT

```{r}
v129_other_2 <- v129_other %>%
  add_flag_CT(vnum="v129_race", old = "U", new = "A")

v129_other_2
```

# compare initial and duplicated calculations

```{r}
compare_subgroup("v129", v129_other_2, 2026)
```


## save data

```{r}
save_data(v129_other_2, "v129_otherdata")
```

