---
title: "NCHS v147 Life Expectancy (LE) - R2026"
author: "GL"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

# about

## Basic information

- Description
Average number of years people are expected to live.

## Raw data access

- NCHS mortality and natality data 2021-2023: the data is requested

- Census population estimates 2021-2023 from:

2021 Census Population Estimates: https://www2.census.gov/programs-surveys/popest/datasets/2020-2021/counties/asrh/cc-est2021-all.csv

2022 Census Population Estimates: https://www2.census.gov/programs-surveys/popest/datasets/2020-2022/counties/asrh/cc-est2022-all.csv

2023 Census Population Estimates: https://www2.census.gov/programs-surveys/popest/datasets/2020-2023/counties/asrh/cc-est2023-alldata.csv


# import packages

```{r}
library(tidyverse)
library(haven)
```

# set up directory

```{r}
# set up directory that holds data needed for calculations
# NCHS mortality data, not allowed to share in GitHub
nchs_mort_dir <- "D:/CHRR/mortality raw"

# NCHS natality data, not allowed to share in GitHub
nchs_nata_dir <- "D:/CHRR/natality raw"

# Census county population estimates data
census_pop_dir <- "D:/CHRR/pop raw"
```


# standard county and state fips

- The final dataset will need to be aligned with standard FIPS codes and to include all counties and states.
```{r }
fips <-  bind_rows(
  read_sas("../inputs/county_fips_with_ct_old.sas7bdat"),
  read_sas( "../inputs/state_fips.sas7bdat")) %>% 
  arrange(statecode, countycode) %>% 
  glimpse()

```

# helpers

## functions used commonly by NCHS measures

- some commonly used functions are in a separate file, `nchs_measure_helpers.R`. 
```{r}
source("../scripts/nchs_measure_helpers.R")
```

## function: process pop data for v147

```{r}
process_pop_v147 <- function(year, by_race_TF = FALSE){
  pop <- read_census_pop(file_name = get_census_csv_file_info(year)$file_name ) %>% 
    get_census_pop_by_agecat_race(get_census_csv_file_info(year)$year_num, by_race = by_race_TF, age_cat_version = "v147")
  
  return(pop)
}
```

## function: process mort data for v147

```{r}
process_mort_v147 <- function(year, by_race_TF = FALSE){
  
  print(get_mort_file_name(year))
  
  mort <- read_mort_data(
    file_path = file.path(nchs_mort_dir, get_mort_file_name(year) ), 
    header = mort_header,
    n_max = Inf) %>% 
    get_mort_data_by_agecat_race(df_fips=fips, by_race = by_race_TF, age_cat_version = "v147")
  
  return(mort)
}

```

# Pop data

## NCHS birth data: age_cat = 0

- Use births data to approximate infant population (age_cat = 0), as Census county-level population estimates use 5-year age groups and do not have age_cat = 0.

- The NCHS birth data was aggregated by state, county, age category, and race/ethnicity group.

```{r}
yrs <- c(2021:2023)

infant_pop <- 
  get_infant_pop(
    yrs, 
    file.path(nchs_nata_dir, "nchs_births_2017_2023.csv"), 
    by_race = FALSE) %>% 
  glimpse()

infant_pop
```

## Census pop data 2021-2023 by age_cat 1-18 

```{r}
pop_agecat_1_18 <- bind_rows(
  process_pop_v147(2021),
  process_pop_v147(2022) %>% 
    fix_CT_pop(process_pop_v147(2021)),
  process_pop_v147(2023) %>% 
    fix_CT_pop(process_pop_v147(2021))
  ) %>% 
  group_by(statecode, countycode, age_cat) %>% 
  summarise(pop = sum(pop, na.rm = TRUE), .groups = "drop") %>% 
  glimpse()

```


## combine age cat 0 with age cat 1-18

```{r}
pop_all <- 
  bind_rows(
    # age_cat = 2 to 18
    pop_agecat_1_18 %>% filter(age_cat!=1),
    
    # age_cat = 0
    fips %>% 
      select(1:2) %>% 
      left_join(infant_pop, by = c('statecode', 'countycode')) %>% 
      mutate(pop = if_else(is.na(pop), 0, pop)) %>% 
      mutate(age_cat = 0),
    
    # fixed age_cat = 1
    pop_agecat_1_18 %>% 
      filter(age_cat==1) %>% 
      bind_rows(infant_pop) %>% 
      pivot_wider(names_from = age_cat, values_from = pop, names_glue = "pop_{age_cat}") %>% 
      mutate(pop = if_else(is.na(pop_0), pop_1, pop_1 - pop_0)) %>% 
      select(-c(pop_0, pop_1)) %>% 
      mutate(age_cat = 1)
    ) %>% 
    arrange(statecode, countycode, age_cat) %>% 
  # remove 02261
    filter(!(statecode == "02" & countycode == "261") ) 

pop_all  
```

## pop <0 

```{r}
pop_all %>% 
  filter(pop<0)
```

## set pop < 0 as 0

```{r}
pop_all <- pop_all %>% 
  mutate(pop = if_else(pop<0, 0, pop))

```


# mort data

## header

```{r}
mort_header <- tribble(
  ~fld_name, ~start,  ~end,
  "state_of_residence", 29,  30,
  "county_of_residence", 35,  37,
  "detail_age", 70, 73,
)
```


## mort 2021-2023

```{r}
mort_all <- bind_rows(
  process_mort_v147(2021),
  process_mort_v147(2022),
  process_mort_v147(2023)
  ) %>% 
  group_by(statecode, countycode, age_cat) %>% 
  summarise(deaths = sum(deaths), .groups = "drop") %>% 
  glimpse()

```

# Life Expectancy (LE) CALCULATION 

## merge mort and pop data

```{r}
merged <- pop_all %>% 
  left_join(mort_all, 
            by = c("statecode", "countycode", "age_cat")) %>% 
  # replace NA with 0 for deaths
  mutate(deaths = if_else(is.na(deaths), 0, deaths)) %>% 
  glimpse()

```

## LE table

-   $x_i$ : Age at Start of Interval (Years)

-   age_interval: Age Interval (Years) 	

-   $n_i$ : Interval Width (Years)

-   $a_i$: Average proportion of age interval lived for those who die	

-   $deaths_i$ :Number Of Deaths In Interval 

-   $pop_i$ : Population Years At Risk 

-   $m_i$  : Death Rate In Age Interval

$$m_i = \frac{death_i}{pop_i}$$

-   $q_i$  : Probability Of Dying In Interval	

$$q_i = \frac{n_i\cdot{ m_i}}{1+(1-a_i)\cdot n_i\cdot m_i}$$

Ref: Chin Long Chiang, On Constructing Current Life Tables, Journal of the American Statistical Association, 1972


-   $p_i$  : Probability of Surviving the Interval

$$p_i = 1-q_i$$

```{r}
helper_tbl <- tibble::tribble(
  ~age_cat, ~age_interval,  ~n_i,    ~a_i,  
   0L,        "0",        1L,   0.1,
   1L,      "1-4",        4L,   0.5,
   2L,      "5-9",        5L,   0.5,
   3L,    "10-14",        5L,   0.5,
   4L,    "15-19",        5L,   0.5,
   5L,    "20-24",        5L,   0.5,
   6L,    "25-29",        5L,   0.5,
   7L,    "30-34",        5L,   0.5,
   8L,    "35-39",        5L,   0.5,
   9L,    "40-44",        5L,   0.5,
  10L,    "45-49",        5L,   0.5,
  11L,    "50-54",        5L,   0.5,
  12L,    "55-59",        5L,   0.5,
  13L,    "60-64",        5L,   0.5,
  14L,    "65-69",        5L,   0.5,
  15L,    "70-74",        5L,   0.5,
  16L,    "75-79",        5L,   0.5,
  17L,    "80-84",        5L,   0.5,
  18L,      "85+",        NA,   0.5,
  )
```


```{r}
le_table <- merged %>% 
  rename(deaths_i = deaths, pop_i = pop) %>% 
  left_join(helper_tbl, by = "age_cat") %>% 
  group_by(statecode, countycode) %>% 
  # calculate rates
  mutate(m_i = deaths_i/pop_i,
         q_i = n_i * m_i / (1 + (1 - a_i) * n_i * m_i),
         p_i = 1 - q_i
    ) %>%  
  mutate(
    q_i = if_else(age_cat == 18, 1, q_i), # fix for xi = 85
    p_i = if_else(age_cat == 18, 0, p_i)  # fix for xi = 85
  ) %>% 
  # Hypothetical Life Table Cohort
  mutate(l_i = accumulate(q_i, ~ .x * (1 - .y), .init = 100000)[-length(q_i)-1],
         d_i = q_i*l_i,
         L_i = (l_i-d_i)*n_i + (d_i * n_i * a_i))  %>%  
  mutate(L_i = if_else(age_cat == 18, l_i/m_i, L_i)) %>% # fix for xi = 85
  mutate(T_i = rev(cumsum(rev(L_i)))) %>%  # reverse cumsum
  mutate(e_i = T_i / l_i)

le_table
```

## calculate variance and confidence interval

-   Calculate var(qi) = (qi2*(1-qi))/Deaths, var(qi) = 0 where Deaths=0

    *   the (binomial) variance of the survival probability 𝑝𝑥 for the 𝑥𝑡ℎ age interval.

$When\space death_i\neq 0$

$$var(q_i)=\frac{q_i^2\cdot(1-q_i)}{death_i}$$

$When\space death_i=0$

$$var(q_i)=0$$

-   Calculate v1 = li2*(((1-ai)*ni+ei+1)2)*var(qi)

$$v1=l_i^2\cdot [(1-a_i)\cdot n_i + e_{i+1}]^2\cdot var(q_i)$$

-   Calculate v2 = v1 + v2i+1, v2w=0

$$v2_i = v1_i +v2_{i+1}$$
v2: Reverse cumulative sum of v1

-   Calculate var(ei) = v2/(li2)

$$var(ei)=\frac{v2}{l_i^2}$$

-   Calculate paw = lw/li

$$paw=\frac{l_w}{l_i}$$

-   Calculate term_adj = paw2/(Mw3*Populationw)

$$ term_{adj} = \frac{paw^2}{M_w^3\cdot Population_w^2}$$

-   Calculate var_adj = var(ei)+term_adj

$$var\_adj = var(e_i)+term_{adj}$$

-   Calculate standard error SE(ei) = sqrt(var_adj)

$$SE(e_i)=\sqrt {var\_adj}$$

-   Calculate CILL and CIUL = ei(+ or -)1.96*SE(ei) 

$$ CILL\space or\space CIUL=e_i \pm SE(e_i) $$

```{r}
le_table_var <- le_table %>% 
  group_by(statecode, countycode) %>%
  mutate(var_qi = case_when(
    deaths_i == 0 ~ 0,
    deaths_i != 0 ~ q_i^2 * (1 - q_i)/deaths_i
  )) %>% 
  mutate(e_i_lead = lead(e_i)) %>% #"next" (lead()) values in e_i; (e_i+1, or move one row up)
  mutate(v1 = l_i^2 * ((1-a_i)*n_i+e_i_lead)^2 * var_qi) %>% 
  select(-e_i_lead) %>% # remove column e_i_lead
  mutate(v2 = rev(cumsum(rev(replace_na(v1, 0))))) %>% 
  mutate(var_ei = v2 / l_i^2) %>% 
  mutate(paw = l_i[length(l_i)]/l_i) %>% 
  mutate(term_adj = paw^2 / (m_i[length(m_i)]^3 * pop_i[length(pop_i)])) %>% 
  mutate(var_adj = var_ei + term_adj) %>% 
  mutate(se = sqrt(var_adj),
         ci = se * 1.96,
         ci_lo = e_i - ci,
         ci_hi = e_i + ci) %>% 
  ungroup()

le_table_var
```


## LE from birth

```{r}
v147 <- le_table_var %>% 
  filter(age_cat == 0) %>% 
  select(statecode, countycode, v147_rawvalue = e_i, v147_cilow = ci_lo, v147_cihigh = ci_hi)

v147  
```

```{r}
summary(v147)

v147 %>% 
  filter(is.infinite(v147_rawvalue))
```


## suppression

-   suppression for counties with pop < 5000

-   suppression 02261??

- County 02013: LE = 146.4, suppressed per verification

```{r}
le_suppress <- le_table %>% 
  select(statecode, countycode, pop_i) %>% 
  group_by(statecode, countycode) %>% 
  summarise(pop = sum(pop_i, na.rm = TRUE), .groups = "drop") %>% 
  mutate(suppress = if_else(pop<5000, 1, 0)) 

le_suppress
```

```{r}
v147_2 <- v147 %>% 
  left_join(le_suppress %>% select(-pop),
            by = c("statecode", "countycode")) %>% 
  mutate(across(starts_with("v147"), ~if_else(suppress == 1, NA, .))) %>% 
  # suppress 02013
  mutate(across(starts_with("v147"), ~if_else(statecode == "02" & countycode == "013", NA, .))) %>% 
  select(-suppress) %>% 
  mutate(across(starts_with("v147"), ~if_else(is.infinite(.), NA, .))) %>% 
  arrange(statecode, countycode)

v147_2
```

```{r}
summary(v147_2)
```

# add flag_CT

- 'A' = data *available* for a CT county
- 'U' = data *unavailable* for a CT county

```{r}
v147_3 <- v147_2 %>% 
  add_flag_CT(vnum = "v147", old="A", new = "U")
```

# Compare to initial calculations 
```{r}
compare("v147", v147_3, year = 2026)
```


# save data
```{r}
save_data(v147_3, "v147")
```
